---
title: "Introducing tidyfun"
author: "Fabian Scheipl"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      fig_width: 12
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introducing tidyfun}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  cache = TRUE,
  fig.width = 8, fig.height = 5.5)

library(tidyverse)
library(refund)
library(ggplot2)
library(patchwork)
library(viridisLite)
theme_set(theme_minimal())
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

try(devtools::load_all("~/fda/tidyfun"))
try(devtools::load_all("~/Work/fda/tidyfun"))

pal_5 = viridis(7)[-(1:2)]
set.seed(1221)
```

## **`tidyfun`**

The goal of **`tidyfun`** is to provide accessible and well-documented software 
that **makes functional data analysis in `R` easy**, specifically:  
data wrangling and exploratory analysis.

**`tidyfun`** provides:  

- new **data types** for representing functional data: **`tfd`** & **`tfb`**

- arithmetic **operators**, descriptive **statistics** and **graphics** functions for such data

- `tidyverse`-verbs for handling functional data **inside** data frames.

# `tf`-Class: Definition

##  `tf`-class

**`tf`** is a new data type for (vectors of) functional data: 

- an abstract superclass for functional data in 2 forms:
    - as (argument, value)-tuples: subclass **`tfd`**, also irregular or sparse
    - or in basis representation: subclass **`tfb`**
    
- basically, a `list` of numeric vectors  
  (... since `list`s work well as columns of data frames ...)

- with additional attributes that define *function-like* behavior:
    - how to **evaluate** the given "functions" for new arguments
    - their **domain** 
    - the **resolution** of the argument values

- `S3` based

## Example Data

Some example data used in this vignette: 

```{r load, echo=FALSE}
dti = with(refund::DTI, 
  data.frame(id = ID, sex = sex, 
    case = factor(ifelse(case, "MS", "control")))) %>% as.tbl %>% 
        mutate(cca = tfd(DTI$cca, seq(0,1, l = 93), signif = 2) %>%
                     tfd(arg = seq(0,1,l = 93)),
               rcst = tfd(DTI$rcst, seq(0, 1, l = 55), signif = 3))
ex_data = dti$cca[1:5, seq(0, 1, l = 93), interpolate = TRUE]
rownames(ex_data) = LETTERS[1:5]
ex = tfd(ex_data, signif = 2)
```

```{r, ex-fig, echo = FALSE}
plot(ex, xlim = c(-0.15, 1), col = pal_5)
text(x = -.1, y = ex[,0.07], labels = names(ex), col = pal_5)
```

```{r}
ex
```

A prepped version of `refund::DTI`:

```{r}
dti
```

## **`tf`** subclass: **`tfd`**

**`tfd`** objects contain "raw" functional data: 

 - represented as a list of **`evaluations`** $f_i(t)|_{t=t'}$ and corresponding **`arg`**ument vector(s) $t'$
 - has a **`domain`**:  the range of valid **`arg`**s.

```{r}
ex %>% evaluations() %>% str
ex %>% arg() %>% str
ex %>% domain() 
```

- each **`tfd`**-vector contains an **`evaluator`** function that defines how to inter-/extrapolate `evaluations` between `arg`s (and remembers results of previous calls)

```{r}
evaluator(ex) %>% str
evaluator(ex) = approx_spline 
```

## **`tf`** subclass: **`tfd`** 

- **`tfd`** has subclasses for regular data with a common grid and irregular or sparse data,
like the `rcst` measurements in our `dti` data.

```{r}
dti$rcst[1:2]
dti$rcst[1:2] %>% arg() %>% str
dti$rcst[1:2] %>% plot(pch = "x", col = viridis(2))
```

## **`tf`** subclass: **`tfb`**

Functional data in basis representation: 

 - represented as a list of **`coefficients`** and a common **`basis_matrix`** of basis function evaluations on a vector of `arg`-values.
 - contains a **`basis`** function that defines how to compute the basis for new **`arg`**s and how to differentiate/integrate.
 - (internal) flavors: **`mgcv`**-spline bases and FPCs (wavelets to be added)
- significant memory and time savings:
```{r}
refund::DTI$cca %>% object.size() %>% print(units = "Kb")
dti$cca %>% object.size() %>% print(units = "Kb")
dti$cca %>% tfb(verbose = FALSE) %>% object.size() %>% print(units = "Kb")
```

## **`tf`** subclass: **`tfb`** spline basis

- accepts all arguments of `mgcv`'s `s()`-syntax 
- either does a penalized fit with (GCV-based) function-specific smoothing or unpenalized.

```{r, message = TRUE}
ex_b = ex %>% tfb(); ex_b[1:2]
ex[1:2] %>% tfb(bs = "tp", k = 55)
```

```{r, eval = FALSE}
ex  %>% plot()
ex_b %>% plot(col = "red")
ex %>% tfb(k = 35, penalized = FALSE) %>% lines(col = "blue")
```
```{r, echo = FALSE, results = 'hide'}
layout(t(1:2))
plot(ex, alpha = 1)
plot(ex_b, col = "red")
lines(tfb(ex, k = 35, penalized = FALSE), col = "blue")
```

## **`tf`** subclass: **`tfb`** FPC-based

- uses either 
    - simple unregularized SVD of the data matrix ("`smooth = FALSE`")
    - or smoothed covariance estimate from `refund::fpca.sc`
- corresponding FPC basis and mean function saved as `tfd`-object
- observed functions are linear combinations of those.

```{r}
(ex %>% tfb_fpc(smooth = FALSE, pve = .999))
(ex %>% tfb_fpc(pve = .95))
```

```{r, eval = FALSE}
ex %>% plot()
ex %>% tfb_fpc(smooth = FALSE, pve = .999) %>% plot(col = "red")
ex %>% tfb_fpc(pve = .95) %>% lines(col = "blue")
```

```{r, echo = FALSE, results = 'hide'}
layout(t(1:2))
plot(ex, alpha = 1)
plot(ex  %>% tfb_fpc(smooth = FALSE, pve = .999), col = "red", ylab = "tfb_fpc(ex)")
lines(ex %>% tfb_fpc(pve = .95), col = "blue")
```

# `tf`-Class: Methods

**`tidyfun`** implements almost all types of operations that are available for conventional
numerical or logical vectors for `tf`-vectors as well, so you can:

### subset & subassign:

```{r}
ex[1:2]
ex[1:2] = ex[2:1]
ex
```

### compare & compute:

```{r, echo = FALSE}
n_ex = names(ex)
ex = unname(ex)
```

```{r}
ex[1] + ex[1] == 2 * ex[1]
log(exp(ex[2])) == ex[2]
ex - (2:-2) != ex 
```

```{r, echo = FALSE}
names(ex) = n_ex
```

### summarize: 

```{r}
c(mean = mean(ex), sd = sd(ex))

depth(ex) ## Modified Band-2 Depth (à la Sun/Genton/Nychka, 2012), others to come.
median(ex) == ex[which.max(depth(ex))]
```

In addition, **`tidyfun`** provides methods specific for functional data:

### evaluate:

`tf`-objects have a special `[`-operator: Its second argument specifies 
`arg`ument values at which to evaluate the functions and has some additional options, 
so it's easy to get point values for `tf` objects: 

```{r, warning  = FALSE}
ex[1:2, seq(0, 1, l = 3)]
ex["B", seq(0, .15, l = 3), interpolate = FALSE]
ex[1:2, seq(0, 1, l = 7), matrix = FALSE] %>% str
```


### (simple, local) smoothing

```{r, echo = FALSE}
layout(t(1:3))
ex %>% plot(alpha = .2, ylab = "lowess")
ex %>% smooth("lowess") %>% lines(col = pal_5)
plot(ex, alpha = .2, ylab = "rolling median (k=5)")
lines(smooth(ex, "rollmedian", k = 5), col = pal_5)
plot(ex, alpha = .2, ylab = "Savitzky-Golay (quartic, 11 steps)")
lines(smooth(ex, "savgol", fl = 11), col = pal_5)
```
```{r, eval  = FALSE}
ex %>% smooth("lowess") %>% plot
ex %>% smooth("rollmedian", k = 5) %>% plot
ex %>% smooth("savgol", fl = 11) %>% plot
```

### differentiate & integrate:

```{r, echo = FALSE}
layout(t(1:3))
plot(ex, col = pal_5)
plot(deriv(smooth(tfd(ex, signif = 4))), col = pal_5, ylab = "deriv(smooth(ex))")
plot(integrate(ex, definite = FALSE), col = pal_5)
```
```{r, eval  = FALSE}
ex %>% plot
ex %>% smooth() %>% deriv() %>% plot
ex %>% integrate(definite = FALSE) %>% plot
```

```{r}
ex %>% integrate()
```

### query

**`tidyfun`** makes it easy to find (ranges of) `arg`uments $t$ satisfying a condition on `value` $f(t)$ (and `arg`ument $t$):

```{r}
ex %>% anywhere(value > .65)
ex[1:2] %>% where(value > .6, "all")
ex[2] %>% where(value > .6, "range")
ex %>% where(value > .6 & arg > .5, "first")
```

### zoom & query

```{r, ex-fig2, echo = FALSE}
plot(ex,  xlim = c(-0.15, 1), col = pal_5, lwd = 2)
text(x = -.1, y = ex[,0.07], labels = names(ex), col = pal_5, cex = 1.5)
lines(median(ex), col = pal_5[3], lwd = 4)
```

```{r}
ex %>% where(value == max(value), "first")
ex[c("A", "D")] %>% zoom(.5, 1) %>% where(value == max(value), "first")
ex %>% zoom(0.2, 0.6) %>% anywhere(value <= median(ex)[, arg])
```

### convert & construct

To & from list, matrix or data frame with `"id"`,`"arg"`,`"value"`-columns:

```{r}
ex_matrix = ex %>% as.matrix(); ex_matrix[1:2, 1:3]
ex_df = ex %>% as.data.frame(); str(ex_df)
ex_matrix[1:2, ] %>% tfd()
tfd(ex_df) == tfd(ex_matrix)
```

# New geoms and stats for functional data for customizable `ggplot2` graphics

### visualize with `base` graphics

```{r ex-fig3}
layout(t(1:2))
plot(ex, type = "spaghetti"); lines(c(median(ex), mean(ex)), col = c(2, 4))
plot(ex, type = "lasagna", col = viridis(50))
```

### visualize with `ggplot2`

**Pasta-themed** `geom`s and plots with for functional data:

- **`geom_spaghetti`** for lines
- **`geom_meatballs`**  for (lines &) points
- **`gglasagna`** with **`order`**-arguments to sort the lasagna layers

To come:

- **`geom_pappardelle`** for functional boxplots
- **`geom_capellini`** for little sparklines / glyphs on maps etc. 

```{r, eval = FALSE}
ggplot(dti, aes(y = cca, colour = case)) +
  geom_spaghetti() + facet_wrap(~ sex)
```

```{r, dti-fig1, echo = FALSE}
ggplot(dti) + 
  geom_spaghetti(aes(y = cca, col = case, alpha = .2 + .4*(case == "control"))) + facet_wrap(~ sex) + scale_alpha(guide = 'none', range = c(.2, .4))
```

```{r, eval = FALSE}
gglasagna(dti, cca, order = integrate(cca, definite = TRUE)) + facet_wrap(~ case)
```
```{r, dti-fig2, echo = FALSE}  
gglasagna(dti, cca, order = integrate(cca, definite = TRUE), arg = seq(0,1, l = 101)) + 
  theme(axis.text.y = element_text(size = 6)) + 
    facet_wrap(~ case, ncol = 2, scales = "free")
```

# Wrangling `tf`s inside data frames

### wrangling `tf`s inside data frames: `dplyr`

**`dplyr`** verbs **`filter`**, **`select`**, **`mutate`**, **`summarize`** work on **`tf`**-columns - e.g.:

```{r, dplyr}
# group-wise functional means:
dti %>% group_by(case, sex) %>% summarize(mean_rcst = mean(rcst, na.rm = TRUE))       %>% ungroup
# which subjects go below cca = .26:
dti %>% filter(anywhere(cca, value < .26))
# center & scale functional data: 
dti %>% 
  mutate(cca = tfb(cca, verbose = FALSE), 
    cca_z = (cca - mean(cca))/sd(cca)) %>% glimpse
```

### wrangling `tf`s inside data frames: `tidyr`

**`tidyfun`** provides `tf_` variants of `tidyr`-verbs to reshape and reformat functional data while keeping it in sync with other covariates:


- `tf_spread:` `tf` $\rightarrow$ columns for each `arg`
- `tf_gather:` columns for each `arg` $\rightarrow$ `tf`

```{r, tidyr}
# spread tf out into columns for each arg
dti_wide = dti %>% tf_spread(cca); dti_wide[, 1:7] %>% glimpse()  

# collect all columns into a single tf-column 
# (... will try to guess arg from column names, name of tf from their prefix)
dti_wide %>% tf_gather(matches("cca_")) %>% glimpse()
```

- `tf_unnest:` `tf` $\rightarrow$ data in long format (`id`, `arg`, `value`)  
- `tf_nest  :` data in long format (`id`, `arg`, `value`)  $\rightarrow$ `tf`

```{r, tidyr2}
# unnest tf by writing 3 loong columns id, arg, value:
# (will try to avoid unnecessary duplication of columns)
dti_long = dti %>% tf_unnest(cca); dti_long %>% glimpse()  
# nest tf by writing 3 loong columns id, arg, value:
dti_long %>% tf_nest(cca_value, .id = cca_id, .arg = cca_arg) %>% glimpse()
```
