---
title: "Data Manipulation"
author: "Jeff Goldsmith"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      fig_width: 12
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Data Manipulation}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.width = 8, fig.height = 5.5)

library(tidyverse)
library(ggplot2)
library(viridisLite)
theme_set(theme_minimal())
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

library("tidyfun")
#try(devtools::load_all("~/fda/tidyfun"))
#try(devtools::load_all("~/Work/fda/tidyfun"))

pal_5 = viridis(7)[-(1:2)]
set.seed(1221)
```

## Data manipulation using **`tidyfun`**

As stated elsewhere, the goal of **`tidyfun`** is to provide accessible and well-documented software that **makes functional data analysis in `R` easy**. In this vignette, we explore some aspects of data manipulation that are possible using **`tidyfun`**, emphasizing compatibility with the `tidyverse`.

Other vignettes have examined the **`tfd`** & **`tfb`** data types, and how to convert common formats for functional data (e.g. matrices, long- and wide-format data frames, `fda` objects) in these new data types. Because our goal is "tidy" data manipulation for functional data analysis, the result of data conversion processes has been a data frame in which a column contains the functional data of interest. This vignette starts from that point.

Throughout, we make use of some visualization tools -- these are explained in more detail in the Visualization vignette.


## Example Data

The data used in this vignette is the `tidyfun::chf_df` dataset. The data contain minute-by-minute observations of log activity counts (stored as a `tfd` vector called `activity`) over seven days for each of 47 subjects with congestive heart failure. In addition to `id` and `activity`, we observe several covariates. 

```{r view_chf}
data(chf_df)

chf_df
```

```{r plot_chf}
chf_df %>% 
  ggplot(aes(y = activity)) + geom_spaghetti(alpha = .05)
```


## Existing `tidyverse` functions

Dataframes using `tidyfun` to store functional observations can be manipulated using tools from `dplyr`, including `select` and `filter`:

```{r}
chf_df %>% 
  select(id, day, activity) %>% 
  filter(day == "Monday") %>% 
  ggplot(aes(y = activity)) + 
  geom_spaghetti(alpha = .05)
```

Operations using `group_by` and `summarize` are allowed:

```{r}
chf_df %>% 
  group_by(day) %>% 
  summarize(mean_act = mean(activity)) %>% 
  ggplot(aes(y = mean_act, color = day)) + 
  geom_spaghetti()
```

One can `mutate` functional observations -- here we exponentiate the log activity counts to obtain original recordings:

```{r}
chf_df %>% 
  mutate(exp_act = exp(activity)) %>% 
  ggplot(aes(y = exp_act)) + 
  geom_spaghetti(alpha = .05)
```

Functions for data manipulation from `tidyr` are also supported. We illustrate by using `spread` to create new `tfd` observations of activity for each day of the week:

```{r}
chf_df %>% 
  select(id, day, activity) %>% 
  spread(key = day, value = activity)
```

(Note that this has made the data less "tidy" and is therefore not generally recommended, but may be useful in some cases).

It's also possible to join datasets based on non-functional keys. To illustrate, we'll first create a pair of datasets:

```{r}
monday_df = filter(chf_df, day == "Monday") %>% 
  select(id, monday_act = activity)
friday_df = filter(chf_df, day == "Friday") %>% 
  select(id, friday_act = activity)
```

These can be joined using the `id` variable as a key (and then tidied using `gather`):

```{r}
left_join(monday_df, friday_df, by = "id") %>% 
  gather(key = day, value = activity, monday_act:friday_act)
```


## New `tidyfun` functions

Some `dplyr` functions are most useful in conjunction with new functions in `tidyfun`. For example, one might use `filter` with `tf_anywhere` to filter based on the values of observed functions:

```{r}
chf_df %>% 
  filter(tf_anywhere(activity, value > 9)) %>% 
  ggplot(aes(y = activity)) + 
  geom_spaghetti()
```

The existing `mutate` function can be combined with several `tidyfun` functions, including `tf_smooth` and `tf_zoom`.

One can smooth existing observations using `tf_smooth`:

```{r}
chf_df %>% 
  filter(id == 1) %>% 
  mutate(smooth_act = tf_smooth(activity)) %>% 
  ggplot(aes(y = smooth_act)) + 
  geom_spaghetti()
```

And one can extract observations over a subset of the full domain using `tf_zoom`:

```{r}
chf_df %>% 
  filter(id == 1) %>% 
  mutate(daytime_act = tf_zoom(activity, 360, 1200)) %>% 
  ggplot(aes(y = daytime_act)) + 
  geom_spaghetti()
```

As we've seen elsewhere, it's possible to convert from `tfd` to `tfb` vectors; this can be done inside a `mutate` statement as part of a data processing pipeline:

```{r}
chf_df %>% 
  mutate(act_basis = tfb(activity)) %>% 
  select(id, gender, act_basis)
```


## Some not-allowed tidying

Unfortunately, based on the existing infrastructure for non-standard data types in `tidyr`, some operations are not possible at this time. The most important of these to note are binds.

Using the datasets defined above, it would ideally be possible to combine these using `bind_rows`, but the following code does not work:

```{r, error = TRUE}
bind_rows(monday_df, friday_df)
```

In many cases, the need for this operation can be avoided through suitable data wrangling prior to converting functional data to a `tf` object.
