---
title: "Visualization"
author: "Jeff Goldsmith, Fabian Scheipl"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      fig_width: 12
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.width = 8, fig.height = 5.5)

library(tidyverse)
library(ggplot2)
library(viridisLite)
theme_set(theme_minimal())
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

library("tidyfun")
#try(devtools::load_all("~/fda/tidyfun"))
#try(devtools::load_all("~/Work/fda/tidyfun"))

pal_5 = viridis(7)[-(1:2)]
set.seed(1221)
```

## Plotting **`tidyfun`** data

The **`tidyfun`** package is designed to facilitate **functional data analysis in `R`**, with particular emphasis on compatibility with the `tidyverse`. In this vignette, we illustrate data visualization using **`tidyfun`**.

Other vignettes have examined the **`tfd`** & **`tfb`** data types; how to convert common formats for functional data into these new data types; and data manipulation using `tf` vectors in data frames. We'll use these in the following, and focus more specifically on visualiation


## Example Data

The data used in this vignette is the `tidyfun::chf_df` dataset. The data contain minute-by-minute observations of log activity counts (stored as a `tfd` vector called `activity`) over seven days for each of 47 subjects with congestive heart failure. In addition to `id` and `activity`, we observe several covariates. 

```{r view_chf}
data(chf_df)

chf_df
```


## Plotting with `ggplot`

`ggplot` is a powerful framework for visualization. In this section, we'll assume some basic familiarity with the package; if you're new to `ggplot`, [this primer](https://rpubs.com/hadley/ggplot-intro) may be helpful.

### `geom_spaghetti` and `geom_meatballs`

One of the most fundamental plots for functional data is the spaghetti plot, which is implemented in `tidyfun` + `ggplot` through `geom_spaghetti`:

```{r plot_chf}
chf_df %>% 
  filter(id == 1) %>% 
  ggplot(aes(y = activity)) + 
  geom_spaghetti()
```

A variant on the spaghetti plot is the meatballs plot, which shows both the "noodles" (i.e. functional observations visualized as curves) and the "meatballs" (i.e. original data visualized as points). 

```{r}
chf_df %>% 
  filter(id == 1, day == "Monday") %>% 
  ggplot(aes(y = activity)) + 
  geom_meatballs()
```

### Using with `ggplot`

The new geoms in `tidyfun` "play nicely" with standard `ggplot` aesthetics and options. 

You can, for example, define the color aesthetic for plots of `tf` variables using other observations:

```{r}
chf_df %>% 
  filter(id %in% 1:5) %>% 
  ggplot(aes(y = activity, color = gender)) + 
  geom_spaghetti(alpha = .2)
```

You can also use facetting:

```{r}
chf_df %>% 
  filter(day %in% c("Monday", "Sunday")) %>% 
  ggplot(aes(y = activity, color = gender)) + 
  geom_spaghetti(alpha = .1) + 
  facet_grid(~ day)
```

Together with data manipulation tools, this can produce useful exploratory analyses:

```{r}
chf_df %>% 
  group_by(gender, day) %>% 
  summarize(mean_act = mean(activity)) %>% 
  mutate(smooth_mean = tfb(mean_act)) %>% 
  filter(day %in% c("Monday", "Sunday")) %>% 
  ggplot(aes(y = smooth_mean, color = gender)) + 
  geom_spaghetti(size = 1.25, alpha = 1) + 
  geom_meatballs(aes(y = mean_act), alpha = .1) + 
  facet_grid(~ day)
```

### `gglasagna`

Lasagna plots are "[a saucy alternative to spaghetti plosts](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2937254/)". They are a variant on a heatmaps which show functional observations in rows and use color to illustrate values taken at different arguments. In `tidyfun`, lasagna plots are implemented through `gglasagna`:

```{r, eval = FALSE}
chf_df %>% 
  filter(day %in% c("Monday", "Sunday")) %>% 
  gglasagna(activity, label = id, order = day)
```


## Plotting with base R

