---
title: "Visualization"
author: "Jeff Goldsmith, Fabian Scheipl"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      fig_width: 7
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.width = 6,
  fig.height = 4,
  dpi = 72,
  fig.retina = 1,
  out.width = "90%"
)

library(tidyverse)
library(viridisLite)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = "viridis_d",
  ggplot2.discrete.fill = "viridis_d"
)

library("tidyfun")
data(chf_df, package = "tidyfun")
data(dti_df, package = "tidyfun")

pal_5 <- viridis(7)[-(1:2)]
set.seed(1221)
```

The **`tidyfun`** package is designed to facilitate **functional data analysis in `R`**, with particular emphasis on compatibility with the **`tidyverse`**. In this vignette, we illustrate data visualization using **`tidyfun`**.

We'll draw on the `tidyfun::dti_df` and the `tidyfun::chf_df` data sets for examples.   

# Plotting with **`ggplot2`**

**`ggplot2`** is a powerful framework for visualization. In this section, we'll assume some basic familiarity with the package; if you're new to **`ggplot2`**, [this primer](https://rpubs.com/hadley/ggplot-intro) may be helpful.

**`tidyfun`** provides `tf_ggplot()` as the primary interface for plotting functional data with **`ggplot2`**. It works just like `ggplot()`, but understands `tf` vectors -- use the `tf` aesthetic to map a `tf` column to the plot, then add standard **`ggplot2`** geoms (`geom_line`, `geom_point`, `geom_ribbon`, etc.).

For heatmaps, use **`gglasagna`**; for sparklines / glyphs on maps, use **`geom_capellini`** (these use their own specialized stats and are called directly without `tf_ggplot`). The older geoms **`geom_spaghetti`**, **`geom_meatballs`**, and **`geom_errorband`** remain available for backward compatibility.

## `tf_ggplot` with standard geoms

One of the most fundamental plots for functional data is the spaghetti plot, which is implemented in **`tidyfun`** and **`ggplot2`** through `tf_ggplot()` + `geom_line()`:

```{r plot_chf}
dti_df[1:10,] |>
  tf_ggplot(aes(tf = cca)) + geom_line(alpha = .3)
```

Adding `geom_point()` to `geom_line()` shows both the curves and the observed data values:

```{r}
dti_df[1:3,] |>
  tf_ggplot(aes(tf = rcst)) + geom_line(alpha = .3) + geom_point(alpha= .3)
```

## Using with other **`ggplot2`** features

`tf_ggplot()` plays nicely with standard **`ggplot2`** aesthetics and options.

You can, for example, define the color aesthetic for plots of `tf` variables using other observations:

```{r}
chf_df |>
  filter(id %in% 1:5) |>
  tf_ggplot(
    aes(tf = tf_smooth(activity, f = .05), # smoothed inputs for clearer viz
        color = gender)) +
  geom_line(alpha = 0.3)
```

... or use facetting:

```{r}
chf_df |>
  filter((id %in% 1:10) & (day %in% c("Mon", "Sun"))) |>
  tf_ggplot(aes(tf = tf_smooth(activity, f = .05), color = gender)) +
  geom_line(alpha = 0.3, lwd = 1) +
  facet_grid(~day)
```

Another example, using the DTI data, is below.

```{r, dti-fig1}
dti_df |>
  tf_ggplot(aes(tf = cca, col = case, alpha = 0.2 + 0.4 * (case == "control"))) +
  geom_line() + facet_wrap(~sex) +
  scale_alpha(guide = "none", range = c(0.2, 0.4))
```

Together with **`tidyfun`**'s tools for functional data wrangling and summary statistics, the integration with **`ggplot2`** can produce useful exploratory analyses,
like the plot below showing group-wise smoothed and unsmoothed mean activity profiles:

```{r}
chf_df |>
  group_by(gender, day) |>
  summarize(mean_act = mean(activity),
            .groups = "drop_last") |>
  mutate(smooth_mean = tfb(mean_act, verbose = FALSE)) |>
  filter(day %in% c("Mon", "Sun")) |>
  tf_ggplot(aes(color = gender)) +
  geom_line(aes(tf = smooth_mean), linewidth = 1.25) +
  geom_line(aes(tf = mean_act), alpha = 0.1) +
  geom_point(aes(tf = mean_act), alpha = 0.1, size = .1) +
  facet_grid(day~.)
```

... or the plot below showing group-wise mean functions +/- twice their pointwise standard errors:
```{r}
dti_df |>
  group_by(sex, case) |>
  summarize(
    mean_cca = mean(tfb(cca, verbose = FALSE)), #pointwise mean function
    sd_cca = sd(tfb(cca, verbose = FALSE)), # pointwise sd function
    .groups = "drop_last"
  ) |>
  group_by(sex, case) |>
  mutate(
    upper_cca = mean_cca + 2 * sd_cca,
    lower_cca = mean_cca - 2 * sd_cca
  ) |>
  tf_ggplot() +
  geom_line(aes(tf = mean_cca, color = sex)) +
  geom_ribbon(aes(tf_ymin = lower_cca, tf_ymax = upper_cca, fill = sex), alpha = 0.3) +
  facet_grid(sex ~ case)
```

## Multivariate functional data (`tf_mv`)

Vector-valued functional data -- functions $f: \mathbb{R} \to \mathbb{R}^d$ stored as `tf_mv` vectors (see the `tf` package and its "Vector-valued functional data" article) -- can be mapped with the `tf` aesthetic just like univariate `tf` columns. `tf_ggplot()` picks a display that mirrors `tf`'s base-R `plot.tf_mv()`:

- for two-component objects (`d == 2`) the default is a **trajectory**: the planar curve $x(t)$ vs $y(t)$, drawn with `geom_path()`;
- otherwise the default is a **facet** display: value-vs-arg curves with one group per curve and component.

We use the built-in `tf::gait` data: hip and knee joint angles for 39 boys across one gait cycle, bundled into a single two-component `tfd_mv`.

```{r mv-setup}
data(gait, package = "tf")
gait_mv <- tfd_mv(list(hip = gait$hip_angle, knee = gait$knee_angle))
gait_df <- tibble(id = factor(seq_along(gait_mv)), mv = gait_mv)
```

The **facet** display shows one panel per component (value versus cycle phase). Pass `type = "facet"`, which unnests the data to a long form with a `.component` column; add `facet_wrap(~ .component)` yourself:

```{r mv-facet}
gait_df |>
  tf_ggplot(aes(tf = mv), type = "facet") +
  geom_line(alpha = 0.3) +
  facet_wrap(~ .component, scales = "free_y")
```

The **trajectory** display (the `d == 2` default) collapses the phase axis and draws each subject's `(hip, knee)` cyclogram -- the characteristic gait "butterfly" loop. **Use `geom_path()`, not `geom_line()`**: `geom_path()` connects points in argument (time) order, while `geom_line()` would re-sort by `x` and mangle any non-monotone curve.

```{r mv-trajectory}
gait_df |>
  tf_ggplot(aes(tf = mv)) +
  geom_path(alpha = 0.3) +
  labs(x = "hip angle", y = "knee angle")
```

Equivalently, you can map two separate univariate `tf` columns (or two components) to the `tf_x` and `tf_y` aesthetics -- handy when the coordinates live in different columns:

```{r mv-txty}
tibble(hip = gait$hip_angle, knee = gait$knee_angle) |>
  tf_ggplot(aes(tf_x = hip, tf_y = knee)) +
  geom_path(alpha = 0.3)
```

`autoplot()` provides one-line shortcuts -- a trajectory for `d == 2` and a faceted display otherwise -- and `autolayer()` returns a single layer you can add to an existing plot:

```{r mv-autoplot}
autoplot(gait_mv)
```

### Colouring trajectories by argument

A trajectory plot hides *where* along the domain each point lies -- the argument (here: the phase of the gait cycle) is not shown on any axis. Every `tf_ggplot()` layer provides a column `.arg` holding the argument value of each evaluation point, so you can map it to the `colour` aesthetic to colour each segment of a curve by its position along the domain. `autoplot()` offers the same via `colour_by_arg = TRUE`:

```{r mv-arg-colour}
gait_df |>
  tf_ggplot(aes(tf = mv, colour = .arg)) +
  geom_path() +
  scale_colour_viridis_c("phase") +
  labs(x = "hip angle", y = "knee angle")

autoplot(gait_mv, colour_by_arg = TRUE)
```

### Phase-plane plots

Phase-plane plots (see `fda::phaseplanePlot()`) display a *univariate* function through two of its derivatives, most commonly the first derivative (velocity) against the second (acceleration), so that cyclic or oscillatory behaviour shows up as loops. `tf_phaseplane()` turns a `tf` vector into a two-component `tf_mv` object of derivatives (components `D1` and `D2` by default; use `order` to pick other derivatives, `0` denoting the function itself) that can be plotted as a trajectory like any other `tf_mv`:

```{r phaseplane}
# derivatives are computed by finite differences, so we use smoothed knee
# angles on a fine grid
knee_smooth <- tfb(gait$knee_angle, k = 12, verbose = FALSE) |>
  tfd(arg = seq(0.025, 0.975, by = 0.005))
knee_pp <- tf_phaseplane(knee_smooth)
knee_pp

tibble(id = factor(seq_along(knee_pp)), pp = knee_pp) |>
  tf_ggplot(aes(tf = pp, colour = .arg)) +
  geom_path(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_colour_viridis_c("phase") +
  labs(x = "knee angle velocity", y = "knee angle acceleration")
```

To mark specific argument values as in `fda::phaseplanePlot()`, evaluate the phase-plane object at those points with `tf_unnest()` (which yields one column per component for `tf_mv` input) and add a text layer with its own data. Here we label the phase-plane of the mean knee angle at every tenth of the gait cycle:

```{r phaseplane-labels}
mean_pp <- tf_phaseplane(mean(knee_smooth))
marks <- tf_unnest(mean_pp, arg = seq(0.1, 0.9, by = 0.1))
tibble(pp = mean_pp) |>
  tf_ggplot(aes(tf = pp)) +
  geom_path() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_label(
    data = marks, aes(x = D1, y = D2, label = arg),
    inherit.aes = FALSE
  ) +
  labs(x = "knee angle velocity", y = "knee angle acceleration")
```

To unnest a `tf_mv` column to a "wide" long table yourself (one value column per component), use `tf_unnest()`:

```{r mv-unnest}
tf_unnest(gait_mv) |> head()
```

## Functional data boxplots with `geom_fboxplot`

"Boxplots" for functional data are implemented in `tidyfun` through `geom_fboxplot()`. They show the deepest function (using modified band depth, `"MBD"`, by default) as a thick median curve, a filled central ribbon spanning the pointwise range of the deepest half of the sample (by default), and outer fence lines obtained by inflating that central envelope by a factor of `coef = 1.5`. Curves that leave the fence envelope anywhere are flagged as outliers and drawn separately.

Like `geom_boxplot()`, the layer follows the usual **`ggplot2`** interface closely. Grouping works naturally through `colour`, `fill`, or `group`, and the same group colour is reused for the ribbon, the outlier curves, and the median if a group colour/fill is mapped; otherwise, the median defaults to black.

```{r}
dti_df |> 
  tf_ggplot(aes(tf = cca, fill = case)) +
  geom_fboxplot(alpha = 0.35) +
  facet_grid(~ sex) + labs(title="MBD-based boxplot")
```

```{r}
dti_df |>
  tf_ggplot(aes(tf = cca, colour = case)) +
  geom_fboxplot(depth = "FM", alpha = 0.3) +
  facet_grid(~ sex) + labs(title="FM-based boxplot")
```

```{r}
dti_df |>
  tf_ggplot(aes(tf = cca, colour = case)) +
  geom_fboxplot(depth = "RPD", alpha = 0.3) +
  facet_grid(~ sex) + labs(title="RPD-based boxplot")
```

The layer also supports irregular functional data directly:
```{r}
tf_ggplot(dti_df, aes(tf = rcst)) + geom_fboxplot()
```

Useful arguments:

```{r}
tf_ggplot(dti_df, aes(tf = rcst)) + 
  geom_fboxplot(alpha = .5)
tf_ggplot(dti_df, aes(tf = rcst)) + 
  geom_fboxplot(alpha = .5, central = .2)
tf_ggplot(dti_df, aes(tf = rcst)) + 
  geom_fboxplot(alpha = .5, central = .2, outliers = FALSE)
tf_ggplot(dti_df, aes(tf = rcst)) +
  geom_fboxplot(orientation = "y", alpha = .3)
``` 


## Heatmaps for functional data: `gglasagna`

Lasagna plots are "[a saucy alternative to spaghetti plots](https://doi.org/10.1097/EDE.0b013e3181e5b06a)". They are a variant on a heatmaps which show functional observations in rows and use color to illustrate values taken at different arguments. 
Especially for large samples or noisy data, lasagna plots can be more informative than spaghetti plots, which can be hard to read when many curves are plotted together. Lasagna plots also allow for easy visual comparison of groups of curves, and the `order` argument in `gglasagna()` allows you to sort the curves by any variable or function of the data, which can help reveal patterns in the data.

In **`tidyfun`**, lasagna plots are implemented through `gglasagna` (as well as a basic `plot` method, see  ?!?). A first example, using the CHF data, is below. 

```{r}
chf_df |>
  filter(day %in% c("Mon", "Sun")) |>
  gglasagna(activity)
```

A somewhat more involved example, demonstrating the `order` argument and taking advantage of facets, is next. 

```{r, dti-fig2}  
dti_df |>
  gglasagna(
    tf = cca,
    order = tf_integrate(cca, definite = TRUE), #order by area under the curve
    arg = seq(0, 1, length.out = 101)
  ) +
  theme(axis.text.y = element_text(size = 6)) +
  facet_wrap(~ case:sex, ncol = 2, scales = "free")
```

## `geom_capellini`

To illustrate `geom_capellini`, we'll start with some data prep for the iconic Canadian Weather data from **`fda`**:

```{r}
canada <- data.frame(
  place = fda::CanadianWeather$place,
  region = fda::CanadianWeather$region,
  lat = fda::CanadianWeather$coordinates[, 1],
  lon = -fda::CanadianWeather$coordinates[, 2]
)

canada$temp <- tfd(t(fda::CanadianWeather$dailyAv[, , 1]), arg = 1:365)
canada$precipl10 <- tfd(t(fda::CanadianWeather$dailyAv[, , 3]), arg = 1:365) |>
  tf_smooth()

canada_map <-
  data.frame(maps::map("world", "Canada", plot = FALSE)[c("x", "y")])
```

Now we can plot a map of Canada with annual temperature averages in red, precipitation in blue: 

```{r}
ggplot(canada, aes(x = lon, y = lat)) +
  geom_capellini(aes(tf = precipl10),
    width = 4, height = 5, colour = "blue",
    line.linetype = 1
  ) +
  geom_capellini(aes(tf = temp),
    width = 4, height = 5, colour = "red",
    line.linetype = 1
  ) +
  geom_path(data = canada_map, aes(x = x, y = y), alpha = 0.1) +
  coord_quickmap()
```

Another general use case for `geom_capellini` is visualizing FPCA decompositions:

```{r, warning=FALSE}
cca_fpc_tbl <- tibble(
  cca = dti_df$cca[1:30],
  cca_fpc = tfb_fpc(cca, pve = .8), 
  fpc_1 = map(coef(cca_fpc), 2) |> unlist(), # 1st PC loading
  fpc_2 = map(coef(cca_fpc), 3) |> unlist() # 2nd PC loading
) 
# rescale FPCs by sqrt of eigenvalues for visualization
cca_fpcs_1_2 <- 
  tf_basis(cca_fpc_tbl$cca_fpc, as_tfd = TRUE)[2:3] * 
    sqrt(attr(cca_fpc_tbl$cca_fpc, "score_variance")[1:2]) 
# scaled eigenfunctions look like this:
tibble(
   eigenfunction = cca_fpcs_1_2,
   FPC = factor(1:2)
) |> tf_ggplot() + 
  geom_line(aes(tf = eigenfunction, col = FPC)) + 
  geom_hline(yintercept = 0)
```

So FPC1 is mostly a horizontal (level) shift, while FPC2 mostly affects the size and direction of the extrema around 0.13 and 0.8.

```{r, warning=FALSE}
ggplot(cca_fpc_tbl[1:40,], aes(x =  fpc_1, y = fpc_2)) +
  geom_point(size = .5, col = viridis(3)[2]) +
  geom_capellini(aes(tf =cca_fpc),width = .01, height = .01, line.linetype = 1) +
  labs(x = "FPC1 score", y = "FPC2 score")
```
Consequently, we can see here that the horizontal axis representing the 1st FPC scores has the average level of functions increasing from left to right (first FPC function is basically a level shift), while the size and direction of the extrema at around 0.13 and particularly 0.8 change along the vertical axis representing the 2nd FPC scores.

# Plotting with base R

**`tidyfun`** includes several extensions of base R graphics, which operate on `tf` vectors. For example, one can use `plot` to create either spaghetti or lasagna plots, and `lines` to add lines to an existing plot:

```{r}
cca <- dti_df$cca |>
  tfd(arg = seq(0, 1, length.out = 93), interpolate = TRUE)

layout(t(1:2))

plot(cca, type = "spaghetti")
lines(c(median(cca), mean = mean(cca)), col = viridis(3)[c(1, 3)])

plot(cca, type = "lasagna", col = viridis(50))
```

These `plot` methods use all the same graphics options and can be edited like other base graphics:
```{r, ex-fig2}
cca_five <- cca[1:5]

cca_five |> plot(xlim = c(-0.15, 1), col = pal_5, lwd = 2)

text(
  x = -0.1, y = cca_five[, 0.07], labels = names(cca_five), col = pal_5, cex = 1
)

median(cca_five) |> lines(col = pal_5[3], lwd = 4)
```

`tf` also defines a `plot` method for `tf_registration` objects (returned by
`tf_register()`) for visualizing function alignment (registration / phase
variability). Current versions of `tf` return a registration result that stores
the aligned curves, inverse warps, and template and can be inspected directly:
```{r}
pinch_reg <- tf::pinch |> tfb() |> #smooth before registration for better results
  tf_register() 
pinch_reg
summary(pinch_reg)
plot(pinch_reg)
```

The registered curves, inverse warping functions, and template can also be
extracted explicitly for custom plots:
```{r}
layout(t(1:3))
plot(tf::pinch[1:5], col = pal_5, lwd = 2, points = FALSE)

plot(tf_inv_warps(pinch_reg)[1:5], col = pal_5, lwd = 2, points = FALSE)
abline(c(0, 1), col = "grey", lty = 2)

plot(tf_aligned(pinch_reg)[1:5], col = pal_5, lwd = 2)
lines(tf_template(pinch_reg), col = "black", lwd = 3, lty= 3)
```
