---
title: "Visualization"
author: "Jeff Goldsmith, Fabian Scheipl"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      fig_width: 12
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.width = 8,
  fig.height = 5.5,
  out.width = "90%"
)

library(tidyverse)
library(viridisLite)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete <- scale_colour_viridis_d
scale_fill_discrete <- scale_fill_viridis_d

library("tidyfun")
data(chf_df, package = "tidyfun")
data(dti_df, package = "tidyfun")

pal_5 <- viridis(7)[-(1:2)]
set.seed(1221)
```

The **`tidyfun`** package is designed to facilitate **functional data analysis in `R`**, with particular emphasis on compatibility with the **`tidyverse`**. In this vignette, we illustrate data visualization using **`tidyfun`**.

We'll draw on the `tidyfun::dti_df` and the `tidyfun::chf_df` data sets for examples.   

# Plotting with **`ggplot2`**

**`ggplot2`** is a powerful framework for visualization. In this section, we'll assume some basic familiarity with the package; if you're new to **`ggplot2`**, [this primer](https://rpubs.com/hadley/ggplot-intro) may be helpful.

**`tidyfun`** provides `tf_ggplot()` as the primary interface for plotting functional data with **`ggplot2`**. It works just like `ggplot()`, but understands `tf` vectors — use the `tf` aesthetic to map a `tf` column to the plot, then add standard **`ggplot2`** geoms (`geom_line`, `geom_point`, `geom_ribbon`, etc.).

For heatmaps, use **`gglasagna`**; for sparklines / glyphs on maps, use **`geom_capellini`** (these use their own specialized stats and are called directly without `tf_ggplot`). The older geoms **`geom_spaghetti`**, **`geom_meatballs`**, and **`geom_errorband`** remain available for backward compatibility.

## `tf_ggplot` with standard geoms

One of the most fundamental plots for functional data is the spaghetti plot, which is implemented in **`tidyfun`** and **`ggplot2`** through `tf_ggplot()` + `geom_line()`:

```{r plot_chf}
dti_df[1:10,] |>
  tf_ggplot(aes(tf = cca)) + geom_line(alpha = .3)
```

Adding `geom_point()` to `geom_line()` shows both the curves and the observed data values:

```{r}
dti_df[1:3,] |>
  tf_ggplot(aes(tf = rcst)) + geom_line(alpha = .3) + geom_point(alpha= .3)
```

## Using with other **`ggplot2`** features

`tf_ggplot()` plays nicely with standard **`ggplot2`** aesthetics and options.

You can, for example, define the color aesthetic for plots of `tf` variables using other observations:

```{r}
chf_df |>
  filter(id %in% 1:5) |>
  tf_ggplot(aes(tf = tf_smooth(activity, f = .05), color = gender)) +
  geom_line(alpha = 0.3)
```

... or use facetting:

```{r}
chf_df |>
  filter((id %in% 1:10) & (day %in% c("Mon", "Sun"))) |>
  tf_ggplot(aes(tf = tf_smooth(activity, f = .05), color = gender)) +
  geom_line(alpha = 0.3) +
  facet_grid(~day)
```

Another example, using the DTI data, is below.

```{r, dti-fig1}
dti_df |>
  tf_ggplot(aes(tf = cca, col = case, alpha = 0.2 + 0.4 * (case == "control"))) +
  geom_line() + facet_wrap(~sex) +
  scale_alpha(guide = "none", range = c(0.2, 0.4))
```

Together with **`tidyfun`**'s tools for functional data wrangling and summary statistics, the integration with **`ggplot2`** can produce useful exploratory analyses,
like the plot below showing group-wise smoothed and unsmoothed mean activity profiles:

```{r}
chf_df |>
  group_by(gender, day) |>
  summarize(mean_act = mean(activity),
            .groups = "drop_last") |>
  mutate(smooth_mean = tfb(mean_act, verbose = FALSE)) |>
  filter(day %in% c("Mon", "Sun")) |>
  tf_ggplot(aes(color = gender)) +
  geom_line(aes(tf = smooth_mean), linewidth = 1.25) +
  geom_line(aes(tf = mean_act), alpha = 0.1) +
  geom_point(aes(tf = mean_act), alpha = 0.1, size = .1) +
  facet_grid(day~.)
```

... or the plot below showing group-wise mean functions +/- twice their pointwise standard errors:
```{r}
dti_df |>
  group_by(sex, case) |>
  summarize(
    mean_cca = mean(tfb(cca, verbose = FALSE)), #pointwise mean function
    sd_cca = sd(tfb(cca, verbose = FALSE)), # pointwise sd function
    .groups = "drop_last"
  ) |>
  group_by(sex, case) |>
  mutate(
    upper_cca = mean_cca + 2 * sd_cca,
    lower_cca = mean_cca - 2 * sd_cca
  ) |>
  tf_ggplot() +
  geom_line(aes(tf = mean_cca, color = sex)) +
  geom_ribbon(aes(tf_ymin = lower_cca, tf_ymax = upper_cca, fill = sex), alpha = 0.3) +
  facet_grid(sex ~ case)
```

## Functional data boxplots with `geom_fboxplot`

"Boxplots" for functional data are implemented in `tidyfun` through `geom_fboxplot()`. They show the deepest function (using modified band depth, `"MBD"`, by default) as a thick median curve, a filled central ribbon spanning the pointwise range of the deepest half of the sample (by default), and outer fence lines obtained by inflating that central envelope by a factor of `coef = 1.5`. Curves that leave the fence envelope anywhere are flagged as outliers and drawn separately.

Like `geom_boxplot()`, the layer follows the usual **`ggplot2`** interface closely. Grouping works naturally through `colour`, `fill`, or `group`, and the same group colour is reused for the ribbon, the outlier curves, and the median if a group colour/fill is mapped; otherwise, the median defaults to black.

```{r}
dti_df |> 
  tf_ggplot(aes(tf = cca, fill = case)) +
  geom_fboxplot(alpha = 0.35) +
  facet_grid(~ sex) + labs(title="MBD-based boxplot")
```

```{r}
dti_df |>
  tf_ggplot(aes(tf = cca, colour = case)) +
  geom_fboxplot(depth = "FM", alpha = 0.3) +
  facet_grid(~ sex) + labs(title="FM-based boxplot")
```

```{r}
dti_df |>
  tf_ggplot(aes(tf = cca, colour = case)) +
  geom_fboxplot(depth = "RPD", alpha = 0.3) +
  facet_grid(~ sex) + labs(title="RPD-based boxplot")
```

The layer also supports irregular functional data directly:
```{r}
tf_ggplot(dti_df, aes(tf = rcst)) + geom_fboxplot()
```

Useful arguments:

```{r}
tf_ggplot(dti_df, aes(tf = rcst)) + 
  geom_fboxplot(alpha = .5)
tf_ggplot(dti_df, aes(tf = rcst)) + 
  geom_fboxplot(alpha = .5, central = .2)
tf_ggplot(dti_df, aes(tf = rcst)) + 
  geom_fboxplot(alpha = .5, central = .2, outliers = FALSE)
tf_ggplot(dti_df, aes(tf = rcst)) +
  geom_fboxplot(orientation = "y", alpha = .3)
``` 


## Heatmaps for functional data: `gglasagna`

Lasagna plots are "[a saucy alternative to spaghetti plots](https://pmc.ncbi.nlm.nih.gov/articles/PMC2937254/)". They are a variant on a heatmaps which show functional observations in rows and use color to illustrate values taken at different arguments. 

In **`tidyfun`**, lasagna plots are implemented through `gglasagna`. A first example, using the CHF data, is below. 

```{r}
chf_df |>
  filter(day %in% c("Mon", "Sun")) |>
  gglasagna(activity)
```

A somewhat more involved example, demonstrating the `order` argument and taking advantage of facets, is next. 

```{r, dti-fig2}  
dti_df |>
  gglasagna(
    tf = cca,
    order = tf_integrate(cca, definite = TRUE),
    arg = seq(0, 1, length.out = 101)
  ) +
  theme(axis.text.y = element_text(size = 6)) +
  facet_wrap(~case, ncol = 2, scales = "free")
```

## `geom_capellini`

To illustrate `geom_capellini`, we'll start with some data prep for the iconic Canadian Weather data from **`fda`**:

```{r}
canada <- data.frame(
  place = fda::CanadianWeather$place,
  region = fda::CanadianWeather$region,
  lat = fda::CanadianWeather$coordinates[, 1],
  lon = -fda::CanadianWeather$coordinates[, 2]
)

canada$temp <- tfd(t(fda::CanadianWeather$dailyAv[, , 1]), arg = 1:365)
canada$precipl10 <- tfd(t(fda::CanadianWeather$dailyAv[, , 3]), arg = 1:365) |>
  tf_smooth()

canada_map <-
  data.frame(maps::map("world", "Canada", plot = FALSE)[c("x", "y")])
```

Now we can plot a map of Canada with annual temperature averages in red, precipitation in blue: 

```{r}
ggplot(canada, aes(x = lon, y = lat)) +
  geom_capellini(aes(tf = precipl10),
    width = 4, height = 5, colour = "blue",
    line.linetype = 1
  ) +
  geom_capellini(aes(tf = temp),
    width = 4, height = 5, colour = "red",
    line.linetype = 1
  ) +
  geom_path(data = canada_map, aes(x = x, y = y), alpha = 0.1) +
  coord_quickmap()
```

Another general use case for `geom_capellini` is visualizing FPCA decompositions:

```{r, warning=FALSE}
cca_fpc_tbl <- tibble(
  cca = dti_df$cca[1:30],
  cca_fpc = tfb_fpc(cca, pve = .8), 
  fpc_1 = map(coef(cca_fpc), 2) |> unlist(), # 1st PC loading
  fpc_2 = map(coef(cca_fpc), 3) |> unlist() # 2nd PC loading
) 
# rescale FPCs by sqrt of eigenvalues for visualization
cca_fpcs_1_2 <- 
  tf_basis(cca_fpc_tbl$cca_fpc, as_tfd = TRUE)[2:3] * 
    sqrt(attr(cca_fpc_tbl$cca_fpc, "score_variance")[1:2]) 
# scaled eigenfunctions look like this:
tibble(
   eigenfunction = cca_fpcs_1_2,
   FPC = factor(1:2)
) |> tf_ggplot() + 
  geom_line(aes(tf = eigenfunction, col = FPC)) + 
  geom_hline(yintercept = 0)
```

So FPC1 is mostly a horizontal (level) shift, while FPC2 mostly affects the size and direction of the extrema around 0.13 and 0.8.

```{r, warning=FALSE}
ggplot(cca_fpc_tbl[1:40,], aes(x =  fpc_1, y = fpc_2)) +
  geom_point(size = .5, col = "red") +
  geom_capellini(aes(tf =cca_fpc),width = .01, height = .01, line.linetype = 1) +
  labs(x = "FPC1 score", y = "FPC2 score")
```
Consequently, we can see here that the horizontal axis representing the 1st FPC scores has the average level of functions increasing from left to right (first FPC function is basically a level shift), while the size and direction of the extrema at around 0.13 and particularly 0.8 change along the vertical axis representing the 2nd FPC scores.

# Plotting with base R

**`tidyfun`** includes several extensions of base R graphics, which operate on `tf` vectors. For example, one can use `plot` to create either spaghetti or lasagna plots, and `lines` to add lines to an existing plot:

```{r}
cca <- dti_df$cca |>
  tfd(arg = seq(0, 1, length.out = 93), interpolate = TRUE)

layout(t(1:2))

plot(cca, type = "spaghetti")
lines(c(median(cca), mean = mean(cca)), col = c(2, 4))

plot(cca, type = "lasagna", col = viridis(50))
```

These `plot` methods use all the same graphics options and can be edited like other base graphics:
```{r, ex-fig2}
cca_five <- cca[1:5]

cca_five |> plot(xlim = c(-0.15, 1), col = pal_5, lwd = 2)

text(
  x = -0.1, y = cca_five[, 0.07], labels = names(cca_five), col = pal_5, cex = 1.5
)

median(cca_five) |> lines(col = pal_5[3], lwd = 4)
```

`tf` also defines a `plot` method for `tf_registration` objects (returned by
`tf_register()`) for visualizing function alignment (registration / phase
variability). Current versions of `tf` return a registration result that stores
the aligned curves, inverse warps, and template and can be inspected directly:
```{r}
cca_reg <- tf_register(
  cca[1:5],
  method = "landmark",
  landmarks = tf_landmarks_extrema(cca[1:5], "max")
)

cca_reg
summary(cca_reg)
plot(cca_reg)
```

The registered curves, inverse warping functions, and template can also be
extracted explicitly for custom plots:
```{r}
layout(t(1:2))

plot(tf_inv_warps(cca_reg), col = pal_5, lwd = 2)
plot(tf_aligned(cca_reg), col = pal_5, lwd = 2)
lines(tf_template(cca_reg), col = "black", lwd = 4)
```
