---
title: "Visualization"
author: "Jeff Goldsmith"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      fig_width: 12
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.width = 8, fig.height = 5.5)

library(tidyverse)
library(ggplot2)
library(viridisLite)
theme_set(theme_minimal())
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

library("tidyfun")
#try(devtools::load_all("~/fda/tidyfun"))
#try(devtools::load_all("~/Work/fda/tidyfun"))

pal_5 = viridis(7)[-(1:2)]
set.seed(1221)
```

## Plotting **`tidyfun`** data

The **`tidyfun`** package is designed to facilitate **functional data analysis in `R`**, with particular emphasis on compatibility with the `tidyverse`. In this vignette, we illustrate data visualization using **`tidyfun`**.

Other vignettes have examined the **`tfd`** & **`tfb`** data types; how to convert common formats for functional data into these new data types; and data manipulation using `tf` vectors in data frames. We'll use these in the following, and focus more specifically on visualiation


## Example Data

The data used in this vignette is the `tidyfun::chf_df` dataset. The data contain minute-by-minute observations of log activity counts (stored as a `tfd` vector called `activity`) over seven days for each of 47 subjects with congestive heart failure. In addition to `id` and `activity`, we observe several covariates. 

```{r view_chf}
data(chf_df)

chf_df
```


## Plotting with `ggplot`

`ggplot` is a powerful framework for visualization. In this section, we'll assume some basic familiarity with the package; if you're new to `ggplot`, [this primer](https://rpubs.com/hadley/ggplot-intro) may be helpful.

### `geom_spaghetti` and `geom_meatballs`

One of the most fundamental plots for functional data is the spaghetti plot, which is implemented in `tidyfun` + `ggplot` through `geom_spaghetti`:

```{r plot_chf}
chf_df %>% 
  filter(id == 1) %>% 
  ggplot(aes(y = activity)) + 
  geom_spaghetti()
```

A variant on the spaghetti plot is the meatballs plot, which shows both the "noodles" (i.e. functional observations visualized as curves) and the "meatballs" (i.e. original data visualized as points). 

```{r}
chf_df %>% 
  filter(id == 1, day == "Monday") %>% 
  ggplot(aes(y = activity)) + 
  geom_meatballs()
```

### Using with `ggplot`

The new geoms in `tidyfun` "play nicely" with standard `ggplot` aesthetics and options. 

You can, for example, define the color aesthetic for plots of `tf` variables using other observations:

```{r}
chf_df %>% 
  filter(id %in% 1:5) %>% 
  ggplot(aes(y = activity, color = gender)) + 
  geom_spaghetti(alpha = .2)
```

You can also use facetting:

```{r}
chf_df %>% 
  filter(day %in% c("Monday", "Sunday")) %>% 
  ggplot(aes(y = activity, color = gender)) + 
  geom_spaghetti(alpha = .1) + 
  facet_grid(~ day)
```

Together with data manipulation tools, this can produce useful exploratory analyses:

```{r}
chf_df %>% 
  group_by(gender, day) %>% 
  summarize(mean_act = mean(activity)) %>% 
  mutate(smooth_mean = tfb(mean_act)) %>% 
  filter(day %in% c("Monday", "Sunday")) %>% 
  ggplot(aes(y = smooth_mean, color = gender)) + 
  geom_spaghetti(size = 1.25, alpha = 1) + 
  geom_meatballs(aes(y = mean_act), alpha = .1) + 
  facet_grid(~ day)
```

### `gglasagna`

Lasagna plots are "[a saucy alternative to spaghetti plosts](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2937254/)". They are a variant on a heatmaps which show functional observations in rows and use color to illustrate values taken at different arguments. In `tidyfun`, lasagna plots are implemented through `gglasagna`:

```{r}
chf_df %>% 
  filter(day %in% c("Monday", "Sunday")) %>% 
  gglasagna(activity)
```





# New geoms and stats for functional data for customizable `ggplot2` graphics

### visualize with `base` graphics




```{r ex-fig3, eval = FALSE}
ex = dti_df$cca

plot(ex, type = "spaghetti")
lines(c(median(ex), mean(ex)), col = c(2, 4))

plot(ex, type = "lasagna", col = viridis(50))
```

### visualize with `ggplot2`

**Pasta-themed** `geom`s and plots with for functional data:

- **`geom_spaghetti`** for lines
- **`geom_meatballs`**  for (lines &) points
- **`gglasagna`** with **`order`**-arguments to sort the lasagna layers
- **`geom_capellini`** for little sparklines / glyphs on maps etc.

```{r, dti-fig1}
ggplot(dti_df) + 
  geom_spaghetti(aes(y = cca, col = case, alpha = .2 + .4*(case == "control"))) +
  facet_wrap(~ sex) + scale_alpha(guide = 'none', range = c(.2, .4))
```

```{r, dti-fig2}  
gglasagna(dti_df, cca, 
  order = tf_integrate(cca, definite = TRUE), arg = seq(0,1, l = 101)) + 
  theme(axis.text.y = element_text(size = 6)) + 
    facet_wrap(~ case, ncol = 2, scales = "free")
```

Some data prep for the iconic Canadian Weather data:
```{r}
canada <- data.frame(
  place = fda::CanadianWeather$place, 
  region = fda::CanadianWeather$region, 
  lat = fda::CanadianWeather$coordinates[,1], 
  lon = -fda::CanadianWeather$coordinates[,2], 
  region = fda::CanadianWeather$region)

canada$temp <- tfd(t(fda::CanadianWeather$dailyAv[,,1]), arg = 1:365)
canada$precipl10 <- tfd(t(fda::CanadianWeather$dailyAv[,,3]), arg = 1:365) %>%
  tf_smooth

canada_map <- 
  data.frame(maps::map("world", "Canada", plot = FALSE)[c("x", "y")]) 
```

Now plot a map of Canada with annual temperature averages in red, 
precipitation in blue: 
```{r}
 ggplot(canada, aes(x = lon, y = lat)) + 
   geom_capellini(aes(tf = precipl10), width = 4, height = 5, colour = "blue", 
     line.linetype = 1) +
   geom_capellini(aes(tf = temp), width = 4, height = 5, colour = "red",
     line.linetype = 1) +
   geom_path(data = canada_map, aes(x = x, y = y), alpha = .1) +
   coord_quickmap()
```





## Plotting with base R

