---
title: "Curve Registration: Practical Guide and Pitfalls"
output:
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Curve Registration: Practical Guide and Pitfalls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
library(tf)
par(las = 1)
alpha_palette <- c(
  "#00000080", "#DF536B80", "#61D04F80", "#2297E680", "#28E2E580",
  "#CD0BBC80", "#F5C71080", "#9E9E9E80", "#00000080", "#DF536B80"
)
have_srvf <- rlang::is_installed("fdasrvf")
have_fda <- rlang::is_installed("fda")
```

## Registration: Template, Warp, Domain

Registration aligns functions in time by reducing phase variation (timing differences - "horizontal" variability) while preserving amplitude variation (shape/height differences - "vertical" variability).

Use registration if:

- Curves have broadly similar shapes that are shifted, stretched or compressed relative to each other
    - and your analysis does not care about these timing differences 
    - or you want to analyze these timing differences separately from shape differences.
- You can interpret both the **target template** and the resulting warps in domain terms.

Do not use registration as a (default) preprocessing step if:

- Curves come from clearly different shape classes/clusters (see below).
- Timing differences are meaningful for your question and/or cannot be viewed in isolation from shape differences.
- The time deformation from registration creates implausible subject-specific timelines.

**Core problem:**

Registration models observed curves $x_i(t)$ as *time-deformed versions of a template* $m(s)$, where the time deformation is given by a monotone warping function \(h_i(s)\):

\[
x_i(t) \approx m(h_i^{-1}(t))
\]

Key interpretation:

- \(h_i\) maps aligned/system time to observed time.
- \(h_i^{-1}\) maps observed time to aligned time.
- Slope of \(h_i^{-1}\) greater than 1 indicates local compression in observed time (events happen faster), slope less than 1 indicates local dilation (events happen slower).

**Core workflow in `tf`:**

```{r, eval=FALSE, echo = TRUE, class.source = "show"}
# estimate h_i(t):
warpings <- tf_register(x, method = "...")
# align/register curves by unwarping with h_i^{-1}(t):
x_registered <- tf_unwarp(x, warpings)
```

### What Is the Template?

Template choice is the most important modeling choice in registration: every estimated warp is relative to that template.

Default template behavior in `tf_register()`:

| Method | Default template behavior | How to override |
|:--|:--|:--|
| `srvf` | Karcher-type mean shape^[The Karcher mean (also called Fréchet mean) generalizes arithmetic means to general spaces. In this case, it is a centroid in the amplitude quotient space (functions modulo reparameterization), computed iteratively using the elastic (Fisher-Rao) distance rather than pointwise averaging.] estimated iteratively by `fdasrvf` | pass `template = ...` |
| `fda` | arithmetic mean curve (estimated iteratively) | pass `template = ...` |
| `affine` | arithmetic mean curve | pass `template = ...` |
| `landmark` | column-wise mean of landmark locations | pass `template_landmarks = ...` |

Practical rules:

1. If you need interpretable warpings relative to a known reference, supply an explicit template.
2. Template and data need to be on the same domain and grid for `srvf` and `fda`.
3. Always *visualize* raw curves and the template together before trusting warps.

### What is a Warping Function?

`tf_register()` returns warps \(h_i\). In practice, you usually inspect inverse warps \(h_i^{-1}\) via `tf_invert(warp)` because those are directly used for alignment: they map observed time points to aligned time points.

- Warping functions close to the identity line imply there is little phase correction.
- Local slopes different from 1 imply local time dilation ($\tfrac{d}{dt} h_i^{-1}(t) < 1$: $s$ increases more slowly than $t$) or compression ($\tfrac{d}{dt} h_i^{-1}(t) > 1$: $s$ increases more quickly than $t$) of the observed function relative to the template function.
- Large departures from identity or frequent slope changes and crossings of the identity line without clear alignment gain indicate over-warping risk.

#### Domain-Preserving vs Non-domain-preserving Warps

This distinction is fundamental for interpretation and edge behavior:

- **Domain-preserving**: warps map domain endpoints to themselves. Registration using the `srvf`, `fda`, and `landmark` methods in `tf` is domain preserving.
- **Non-domain-preserving**: warps can shift or scale outside (or into a subset of) the observed domain. Only available with the `affine` method in `tf`.

Implications:

1. Non-domain-preserving warps can produce boundary `NA`s after `tf_unwarp()` because parts of the aligned timeline have no observed support.
2. Domain-preserving warps avoid these boundary losses but can still overfit timing.
3. Method choice is therefore not only about alignment quality; it is also about what time transformation is scientifically admissible.

The figure below illustrates how warping functions compose with a template to produce shifted/deformed curves. Note that this shows the *forward simulation* view, $m(h(s))$: applying a warp to generate an observed curve from the template. In practice, registration works in the *inverse* direction — estimating $h_i^{-1}$ to map an observed curve back to system time (i.e., to *align* it to a template).

```{r warp-illustration, echo = FALSE, fig.width = 8, fig.height = 8}
# Synthetic illustration of different warp types
s <- seq(0, 1, length.out = 201)

# Template: 
template_synth <- tfd(exp(-((s - 0.45)^2) / (2 * 0.08^2)) + cos(2*pi*s) + s, arg = s)

# Warp 1: affine shift + scale (not domain-preserving — extends beyond [0,1])
h_shift <- tfd(0.8 * s + 0.12, arg = s)

# Warp 2: smooth diffeomorphism (domain-preserving)
h_smooth_raw <- s - 0.25 * sin(pi * s)
h_smooth <- tfd(
  (h_smooth_raw - h_smooth_raw[1]) / (h_smooth_raw[length(s)] - h_smooth_raw[1]),
  arg = s)

# Warp 3: wiggly diffeomorphism (domain-preserving)
h_wiggly_raw <- .1 * pbeta(s, .1, .1) + .5 * pbeta(s, 25, 25) + 
  .2 * pbeta(s, 4, 20) + .2 * pbeta(s, 100, 10)
h_wiggly <- tfd(h_wiggly_raw, arg = s)  

# Compose template with warps: m(h(s))
warped_shift <- tf_warp(template_synth, h_shift) |> suppressWarnings()
warped_smooth <- tf_warp(template_synth, h_smooth)
warped_wiggly <- tf_warp(template_synth, h_wiggly)

warp_cols <- c("#DF536B", "#2297E6", "#61D04F")

layout(matrix(1:9, 3, 3))
par(mar = c(4, 4, 3, 1))

warp_names <- c("Affine (shift & scale)", "Smooth elastic", "Wiggly elastic")
warps <- list(h_shift, h_smooth, h_wiggly)
warped <- list(warped_shift, warped_smooth, warped_wiggly)

# Column 1: blank, template, blank
plot.new()
plot(template_synth, lwd = 2.5,
     main = "Template m(s)", xlab = "s", ylab = "m(s)")
plot.new()

# Column 2: one warp per row
for (i in 1:3) {
  plot(s, s, type = "l", lty = 3, lwd = 1,
       main = paste0("h(s): ", warp_names[i]),
       xlab = "s", ylab = "h(s)", ylim = c(-0.05, 1.15))
  lines(warps[[i]], col = warp_cols[i], lwd = 2.5)
}

# Column 3: template (dashed) + warped curve per row
for (i in 1:3) {
  plot(template_synth, lwd = 2, lty = 3,
       main = paste0("m(h(s)): ", warp_names[i]),
       xlab = "s", ylab = "m(h(s))")
  lines(warped[[i]], col = warp_cols[i], lwd = 2.5)
}
```
- Note that the affine shift warp is NOT domain-preserving (it extends beyond [0,1]), while the smooth and wiggly elastic warps ARE domain-preserving (they start at 0 and end at 1).
- The *flatter* the warping function, the more it compresses time and the *steeper* the more it dilates the reference "system" time of the template.
- Where the warping function is above the identity line, the template is *ahead* of the warped function, where it is below, the template is *behind* the warped function.

## Example: Shift Registration with Explicit Template 

This minimal example uses simple shift registration, i.e. the warping model assumes $h_i(s) = s + b_i$ (no elastic deformation of the domain, simply shifts the functions' arguments). 
We pass an explicit template to make the target shape transparent:

```{r quickstart, class.source = "fold-show"}
pinch_small <- pinch[1:10]
template_affine <- pinch[7] |> tf_smooth(f= .2)

pinch_warp_shift <- tf_register(
  pinch_small,
  method = "affine",
  template = template_affine,
  type = "shift"
)
pinch_small_reg <- tf_unwarp(pinch_small, pinch_warp_shift)
```

```{r quickstart-plot, echo = FALSE, class.source = "fold-show", fig.width = 8, fig.height = 3}
layout(t(1:3))
plot(
  pinch_small,
  main = "Original Curves",
  xlab = "Observed Time t [sec]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
lines(template_affine, lwd = 2, lty = 3, col = rgb(0,0,.5))
plot(
  tf_invert(pinch_warp_shift),
  points = FALSE,
  main = "Inverse Warping Functions",
  xlab = "Observed Time t [sec]",
  ylab = "Registered Time s [sec]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3)
plot(
  pinch_small_reg,
  main = "Registered Curves",
  xlab = "Registered Time s [sec]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette, 
  points = FALSE
)
```

Quick success checks:

- The template is plausible for the population you are aligning.
- Major features (e.g., peaks) are more aligned after registration.
- Warps are monotone and not excessively bent for this problem.
- Boundary `NA`s are limited and scientifically acceptable.
- Resulting timing shifts are domain-plausible.

## Choosing a Registration Method

| Method | Use when | Avoid when | Template handling | Key arguments | Input grid | Typical failure signs | First fallback |
|:--|:--|:--|:--|:--|:--|:--|:--|
| `srvf` (default) | Smooth, non-linear timing differences with shared shape family | Very noisy data, sparse grids, highly heterogeneous shapes | Implicit Karcher-type mean unless `template` supplied | `method = "srvf"`, `lambda` | regular only | Unstable warps, over-warping, inconsistent reruns | Use/increase penalization; Smooth more and/or evaluate on finer grids; try `affine` or `landmark` |
| `fda` | You want continuous-criterion alignment and can tune criterion | Complex phase-amplitude interactions or severe irregularity | Arithmetic mean unless `template` supplied | `method = "fda"`, `crit = 1/2`, `nbasis`, `lambda` | regular only | Sensitive to criterion choice, weak alignment gains | Change criterion, basis dimension and/or amount of penalization; Try `srvf` or `affine` |
| `affine` | Mostly shift/scale timing variability; need interpretable warps | Strong non-linear timing deformation | Arithmetic mean unless `template` supplied | `method = "affine"`, `type`, bounds | regular + **irregular** | Residual misalignment of local features, boundary `NA`s under stronger shifts/scales | Set stricter/looser bounds; Upgrade to `srvf` |
| `landmark` | Reliable, repeated landmarks are available across curves | Ambiguous/noisy landmarks or mismatched counts | Uses `template_landmarks` (default: column means) | `method = "landmark"`, `landmarks`, `template_landmarks` | regular + **irregular** | Forced/broken alignments from bad landmarks | Check appropriateness of landmarks; Use `srvf` or `affine` |

The `srvf` and `fda` methods require regular grids (`tfd_reg` or `tfb`). The `affine` and `landmark` methods also accept irregular grids (`tfd_irreg`).

Practical default:

1. Start with `affine` to get a simple baseline.
2. Move to `srvf` if you need non-linear timing flexibility.
3. Use `landmark` when domain landmarks are trusted and central.

See [Theoretical Background] below for method-specific mathematical details.

## Pitfalls and Likely Failure Modes

#### Structured shape variation

**Issue:** Registered curves still form distinct shape clusters.

**Remedies:** Stratify first; register within clusters; compare cluster-specific results.

Registration assumes a single shared template. If curves come from different shape families, a single registration will try to compromise and may distort curves from both groups.

#### Noise dominates phase signal

**Issue:** Warps become jagged, wiggly or highly variable across near-identical curves.

**Remedies:** Smooth inputs (more) first; reduce basis flexibility; compare before/after sensitivity.

In our benchmarks, SRVF and FDA tended to degrade more gracefully under noise than affine and landmark methods. Pre-smoothing inputs with `tf_smooth()` before registration can help all methods.

#### Sparse or irregular grids

**Issue:** Registration quality may be affected strongly by interpolation grid choices.

**Remedies:** Re-evaluate on a fine, common grid; run sensitivity checks on grid density.

The `srvf` and `fda` methods require regular grids (`tfd_reg`); interpolate sparse or irregular data to a common regular grid first. The `affine` and `landmark` methods also accept irregular grids (`tfd_irreg`). Grid density can affect results — try at least two grid resolutions to check stability.

#### Boundary and overlap artifacts

**Issue:** Large `NA` regions after unwarping, especially for aggressive affine shifts/scales.

**Remedies:** Inspect boundary behavior explicitly; narrow affine bounds via `shift_range`/`scale_range`.

This issue is specific to the `affine` method, which is the only non-domain-preserving method in `tf`. The `srvf`, `fda`, and `landmark` methods preserve domain endpoints by construction and do not produce boundary `NA`s.

#### Over-warping

**Issue:** Extreme timing distortions with little gain in feature alignment.

**Remedies:** Prefer simpler methods (`affine`), reduce flexibility, or avoid registration for that subset.

This is primarily a risk with flexible methods (`srvf`, `fda`). Affine warps are constrained by design and rarely over-warp. Comparing results across methods can help detect over-warping: if a flexible method's warps are much more variable than an affine baseline without a clear improvement in alignment, over-warping may be occurring.

## Practical Diagnostics Workflow

After any registration run:

1. Plot original curves together with the template you used.
2. Plot registered curves and compare key features to the template.
3. Plot inverse warps and check for plausible deformation.
4. If possible, quantify alignment targets (e.g., spread of peak locations; eigenvalue spectral gap or decay speed of FPCs pre- and post-registration).
5. Check boundary support (`NA` rates) for affine (non-domain-preserving) warps.
6. Re-run with at least one alternative method.
7. Re-run after modest smoothing/grid change; check stability.

We demonstrate this workflow below using the `pinch` data with affine shift registration.

```{r diagnostics, class.source = "fold-show"}
x <- pinch[1:10] 

# Register with affine shift warps:
warp_aff <- tf_register(x, method = "affine", type = "shift")
x_aff <- tf_unwarp(x, warp_aff)

# Check 1: Are global peak locations more concentrated after alignment?
peak_before <- tf_where(x, value == max(value)) |> as.numeric()
peak_after_aff <- tf_where(x_aff, value == max(value)) |> as.numeric()

data.frame(
  metric = c("sd_peak_before", "sd_peak_after_affine"),
  value = c(sd(peak_before), sd(peak_after_aff))
)

# Check 2: How much boundary support is lost?
# For shift warps h(t) = t + delta, the shift is readable from the warp directly:
shifts <- as.numeric(warp_aff[, tf_domain(x)[1]]) - tf_domain(x)[1]
# Proportion of domain lost to boundary NAs (exact for pure shifts):
na_prop_aff <- mean(abs(shifts)) / diff(tf_domain(x))
na_prop_aff
```

```{r diagnostics-plot, fig.width = 8, fig.height = 3}
layout(t(1:4))
# Original curves:
plot(x, main = "Original",
     col = alpha_palette, lwd = 1.5)
# Inverse warps — should be close to identity for mild shifts:
plot(tf_invert(warp_aff), points = FALSE,
     main = "Inverse Warps",
     xlab = "Observed Time", ylab = "Registered Time",
     col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
# Registered curves:
plot(x_aff, main = "Registered",
     col = alpha_palette, lwd = 1.5, points = FALSE)
# Peak locations before vs after:
plot(peak_before, rep(2, length(peak_before)),
     xlim = tf_domain(x), ylim = c(0.5, 2.5),
     pch = 4, cex = 1.5, col = alpha_palette,
     xlab = "Time", ylab = "", yaxt = "n",
     main = "Peak Locations")
points(peak_after_aff, rep(1, length(peak_after_aff)),
       pch = 16, cex = 1.5, col = alpha_palette)
axis(2, at = c(1, 2), labels = c("After", "Before"), las = 1)
```

```{r diagnostics-optional-srvf, eval = have_srvf, include = have_srvf, class.source = "fold-show"}
# Compare with SRVF (non-linear warps):
warp_srvf <- tf_register(x, method = "srvf")
x_srvf <- tf_unwarp(x, warp_srvf)
peak_after_srvf <- tf_where(x_srvf, value == max(value)) |> as.numeric()

data.frame(
  metric = c("sd_peak_after_affine", "sd_peak_after_srvf"),
  value = c(sd(peak_after_aff), sd(peak_after_srvf))
)
```

```{r diagnostics-optional-srvf-plot, eval = have_srvf, include = have_srvf, fig.width = 8, fig.height = 3}
layout(t(1:4))
plot(x, main = "Original",
     col = alpha_palette, lwd = 1.5)
plot(tf_invert(warp_srvf), points = FALSE,
     main = "SRVF Inverse Warps",
     xlab = "Observed Time", ylab = "Registered Time",
     col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(x_srvf, main = "SRVF Registered",
     col = alpha_palette, lwd = 1.5, points = FALSE)
plot(peak_before, rep(2, length(peak_before)),
     xlim = tf_domain(x), ylim = c(0.5, 2.5),
     pch = 4, cex = 1.5, col = alpha_palette,
     xlab = "Time", ylab = "", yaxt = "n",
     main = "Peak Locations")
points(peak_after_srvf, rep(1, length(peak_after_srvf)),
       pch = 16, cex = 1.5, col = alpha_palette)
axis(2, at = c(1, 2), labels = c("After", "Before"), las = 1)
```

If different methods disagree strongly and diagnostics are unstable, treat conclusions as low confidence until you resolve data representation and model-choice sensitivity.

## Theoretical Background 101

Problem is to find an *optimal* deformation of the function's domain:

\[
\text{Find } h_i(s) \text{ so that }  d\left(x_i(h_i(s)), m(s)\right) \to \min
\]

- \(x_i(t)\): observed curve for subject \(i\) at observed times \(t\).
- \(m(s)\): the "template" function you register against.
- \(h_i(s)\): warping function mapping aligned time \(s\) ("system time")  to observed time \(t\) ("wallclock time") for the $i$-th curve. These are returned by `tf_register()`, and registration methods differ in the representations and constraints for these (see below).
- \(h_i^{-1}(t)\): inverse warping (observed time $\to$ system time) used by `tf_unwarp()` to align observed curves.  
- $d(\cdot, \cdot)$: distance or dissimilarity to minimize (e.g., $L_2$ distance of derivatives $\int (x'_i(h_i(s))h'_i(s) - m'(s))^2 ds)$).

Key references: [Marron et al. (2015)](https://doi.org/10.1214/15-EJS1030) provide an overview of amplitude and phase variation in FDA. Ramsay & Silverman (2005, *Functional Data Analysis*, Ch. 7--8) cover landmark and continuous registration. Srivastava & Klassen (2016, *Functional and Shape Data Analysis*) develop the elastic/SRVF framework in depth.

### SRVF Framework

Represents each curve via its *square root velocity function* (SRVF), $q(t) = \dot{x}(t) / \sqrt{|\dot{x}(t)|}$, and aligns curves by minimizing $L_2$ distances between (aligned) SRVFs. This corresponds to minimizing an *elastic* distance metric between functions modulo reparameterization (i.e. under "warping") and transforms the non-linear alignment problem into a simpler optimization on a Hilbert sphere^[more specifically: the positive orthant of such a sphere, which is the space of SRVFs of warping functions]. See [Srivastava et al. (2011)](https://doi.org/10.48550/arXiv.1103.3817) and [Tucker et al. (2013)](https://doi.org/10.1016/j.csda.2012.12.001) for the `fdasrvf` implementation. 

- **Optimization:** Dynamic programming over the space of diffeomorphisms^[smooth, monotone, one-to-one functions, basically.] (via `fdasrvf`).
- **Template:** Karcher mean on the shape manifold (iterative; see footnote in the [template table][What Is the Template?] above), or user-supplied.
- **Domain-preserving:** Yes — warps satisfy $h(t_\min) = t_\min$ and $h(t_\max) = t_\max$.
- **Key arguments:** `method = "srvf"`. Pass `template` to override the Karcher mean. Control warp flexibility via `lambda` (default is `0` for unrestricted/unpenalized warps) and `penalty_method` (`"roughness"` (default), `"geodesic"`, or `"norm"`).
- **Strengths:** In our benchmarks, SRVF tended to be the best general-purpose method under both simple and complex phase and amplitude variability. Makes no assumptions about warping function shape (except domain preservation and monotonicity).
- **Weaknesses:** Can be sensitive to noise and details of curve representation as well as penalization. Shows the most variable performance under template estimation compared to other methods.

### FDA (Continuous Registration Criterion)

Estimates smooth monotone warping functions by maximizing the integrated cross-correlation between the derivative of each registered curve and the first eigenfunction of the derivatives' covariance, which represents the dominant common shape of the sample (criterion `crit=2`, the default) or by minimizing the squared differences between (aligned) function and template (criterion `crit=1`, not recommended). Warping functions are monotone B-spline functions using `fda::register.fd()`. See [Ramsay & Li (1998)](https://doi.org/10.1111/1467-9868.00129) and Ramsay & Silverman (2005, *Functional Data Analysis*, Ch. 8).

- **Optimization:** Continuous registration criteria  with monotone spline warps.
- **Template:** Arithmetic mean, re-estimated over `max_iter` iterations (default 3), or user-supplied.
- **Domain-preserving:** Yes.
- **Key arguments:** `method = "fda"`. Use `crit = 2` (default) to maximize the proportion of variance explained by the first eigenfunction of the registered sample, or `crit = 1` to minimize integrated squared differences. Control warp flexibility via B-spline basis dimension of the warping functions `nbasis` (default is `6`) and their wiggliness via penalty parameter `lambda` (default is `0` for no penalization). In our experience, `crit = 1` without penalization tends to be considerably less reliable than `crit = 2` or penalized variants — the unpenalized L2 criterion can produce strongly distorted warps.
- **Strengths:** Tends to be the most stable method under template estimation. 
- **Weaknesses:** Results are sensitive to the criterion choice and the spline basis configuration.

### Affine (Shift / Scale)

Models warps as linear functions $h(s) = a \cdot s + b$, with three sub-types:

- `type = "shift"` (default): $a = 1$, only horizontal translation ($b$ free).
- `type = "scale"`: only uniform time scaling ($a$ free, $b$ derived from centering).
- `type = "shift_scale"`: both parameters free.

Each curve is aligned independently via bounded L-BFGS-B optimization of the L2 distance to the template. See Ramsay & Silverman (2005, Ch. 7) and [Wang & Gasser (1997)](https://doi.org/10.1214/aos/1069362747) for context on shift/scale alignment models.

- **Template:** Arithmetic mean, or user-supplied.
- **Domain-preserving:** **No** — warps can shift or scale outside the observed domain, producing boundary `NA`s after `tf_unwarp()`.
- **Input grid:** Supports both regular and irregular grids (`tfd_reg` and `tfd_irreg`).
- **Key arguments:** `method = "affine"`, `type`, `shift_range` and `scale_range` to set upper and lower limits for time shifts $b_i$ and time scales $a_i$.
- **Strengths:** Very fast (typically an order of magnitude or more faster than SRVF), highly interpretable parameters.
- **Weaknesses:** Cannot capture non-linear timing deformation. Results are often shortened functions with substantial leading and/or trailing NAs (under strong shifts/scales).

### Landmark

Constructs piecewise linear warps by mapping user-specified landmark positions to target positions. No continuous optimization is performed — the warp is fully determined by the landmark correspondence. See [Kneip & Gasser (1992)](https://doi.org/10.1214/aos/1176348794) and Ramsay & Silverman (2005, Ch. 7).

- **Template:** Defined by column-wise means of the landmark time points matrix, or user-supplied `template_landmarks`.
- **Domain-preserving:** Yes — domain endpoints are used as boundary anchors.
- **Input grid:** Supports both regular and irregular grids (`tfd_reg` and `tfd_irreg`).
- **Key arguments:** `method = "landmark"`, `landmarks` (required: $n \times k$ matrix of landmark positions), `template_landmarks` (optional target positions).
- **Strengths:** Fast and deterministic. In our benchmarks, landmark registration was most robust to datasets with outlier contamination.
- **Weaknesses:** Requires the ability to identify reliable, consistent landmarks across all curves. Fragile when features are ambiguous and often too simplistic if curves have few identifiable features (e.g., only one peak per curve).
- **Landmark detection:** `tf_landmarks_extrema()` can automatically detect local maxima, minima, or zero crossings and cluster them across curves - usually better to pre-smooth noisy inputs with `tf_smooth()` or `tfb` before calling this function. Results will typically be better for user-defined landmarks that are based on domain knowledge.

## Worked Examples

### Pinch Data: Cross-method Comparison

```{r pinch-compare, class.source = "fold-show"}
pinch_small <- pinch[1:10]

warp_aff <- tf_register(pinch_small, method = "affine", type = "shift_scale")
reg_aff <- tf_unwarp(pinch_small, warp_aff)

# tf_landmarks_extrema needs smoothed inputs, 
#     otherwise it tends to detect lots of spurious features:
pinch_small |> tf_landmarks_extrema(which = "max") |> head()
#     ... so in this case, we simply use the global maximum for each curve:
(peak_locs <- pinch_small |> tf_where(value == max(value)) |> unlist() |> as.matrix())
warp_lm <- tf_register(pinch_small, method = "landmark", landmarks = peak_locs)
reg_lm <- tf_unwarp(pinch_small, warp_lm)

# ... but using more than one peak location as the only landmark will get us 
# better alignment here - use times where curve first & last exceeds 3 as well:
pinch_landmarks  <- cbind(
  start = pinch_small |> tf_where(value > 3, return = "first") |> unlist(),
  peak  = peak_locs,
  end   = pinch_small |> tf_where(value > 3, return = "last") |> unlist()
)
pinch_landmarks |> head()
warp_lm2 <- tf_register(pinch_small, method = "landmark", landmarks = pinch_landmarks)
reg_lm2 <- tf_unwarp(pinch_small, warp_lm2)
```


```{r pinch-compare-plot, fig.width = 8, fig.height = 8}
layout(t(matrix(1:12, 4, 3)))
plot.new() 
plot(tf_invert(warp_aff), main = "Affine Inverse Warps", ylab = "",
     points = FALSE, col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(reg_aff, main = "Affine Registered", 
     col = alpha_palette, lwd = 1.5, points = FALSE)
plot(reg_aff, main = "Affine Registered", 
     type = "lasagna", col = hcl.colors(12, rev = TRUE))

plot(pinch_small, main = "Original", 
     col = alpha_palette, lwd = 1.5)
plot(tf_invert(warp_lm), main = "Landmark Inverse Warps \n (Peak only)", ylab = "",
     points = FALSE, col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(reg_lm, main = "Landmark Registered", col = alpha_palette, lwd = 1.5)
plot(reg_lm, main = "Landmark Registered", type = "lasagna", col = hcl.colors(12, rev = TRUE))

plot.new()
plot(tf_invert(warp_lm2), main = "Landmark Inverse Warps \n (Start + Peak + End)", ylab = "",
     points = FALSE, col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(reg_lm2, main = "Landmark Registered", col = alpha_palette, lwd = 1.5)
plot(reg_lm2, main = "Landmark Registered", type = "lasagna", col = hcl.colors(12, rev = TRUE))
```

Note that affine registration with `shift+scale` warps produces some boundary `NA`s because the warps are not domain-preserving, but it does a good job aligning the peaks. Landmark registration with only the peak locations does a decent job aligning the peaks without producing `NA`s, but it does not align the start and end of the curves well. Landmark registration with start + peak + end landmarks does a good job aligning all three features without producing `NA`s.

```{r pinch-compare-srvf-fda, eval = have_srvf && have_fda, class.source = "fold-show"}
warp_srvf <- tf_register(pinch_small, method = "srvf")
reg_srvf <- tf_unwarp(pinch_small, warp_srvf)
warp_fda_unpen <- tf_register(pinch_small, method = "fda")
reg_fda_unpen <- tf_unwarp(pinch_small, warp_fda_unpen)
warp_fda_pen <- tf_register(pinch_small, method = "fda", lambda = 1e-4)
reg_fda_pen <- tf_unwarp(pinch_small, warp_fda_pen)
```

SRVF and FDA both use non-linear warps and tend to be able to align more complex phase variation than the `affine` method, but their warps are more complex and potentially harder to interpret, and registration can go off the rails more easily (as we can see here with the unpenalized `fda`-registration...).

```{r pinch-compare-srvf-fda-plot, eval = have_srvf && have_fda, fig.width = 8, fig.height = 8}
layout(t(matrix(1:12, 4, 3)))

plot.new()
plot(tf_invert(warp_srvf), main = "SRVF Inverse Warps", ylab = "",
     points = FALSE, col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(reg_srvf, main = "SRVF Registered",
     col = alpha_palette, lwd = 1.5, points = FALSE)
plot(reg_srvf, main = "SRVF Registered",
     type = "lasagna", col = hcl.colors(12, rev = TRUE))

plot(pinch_small, main = "Original",
     col = alpha_palette, lwd = 1.5)
plot(tf_invert(warp_fda_unpen), main = "FDA Inverse Warps (unpen.)", ylab = "",
     points = FALSE, col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(reg_fda_unpen, main = "FDA Registered (unpen.)",
     col = alpha_palette, lwd = 1.5, points = FALSE)
plot(reg_fda_unpen, main = "FDA Registered (unpen.)",
     type = "lasagna", col = hcl.colors(12, rev = TRUE))

plot.new()
plot(tf_invert(warp_fda_pen), main = expression("FDA Inverse Warps (" * lambda * " = 1e-4)"),
     ylab = "", points = FALSE, col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
plot(reg_fda_pen, main = expression("FDA Registered (" * lambda * " = 1e-4)"),
     col = alpha_palette, lwd = 1.5, points = FALSE)
plot(reg_fda_pen, main = expression("FDA Registered (" * lambda * " = 1e-4)"),
     type = "lasagna", col = hcl.colors(12, rev = TRUE))
```

### Growth Data: Deeper Real-data Workflow

The Berkeley growth data contains height measurements for 39 boys and 54 girls aged 1--18. Growth velocity curves (first derivatives of height) show a prominent pubertal growth spurt whose timing varies substantially between individuals — a natural target for registration. We use the subset of girls^[.. because aligning female *and* male growth curves to the same template does not make sense -- they show different patterns!] from this dataset to illustrate how data representation, penalization, and landmark choice affect registration quality.

```{r growth-prep, class.source = "fold-show"}
growth <- tf::growth |> dplyr::filter(gender == "female")

# Raw velocity via finite differences — noisy, only 30 midpoints from 31 measurements:
growth$raw_vel <- tf_derive(growth$height)

# Smooth velocity via spline representation on a much denser grid, then derive analytically:
growth$smooth_vel <- tfb(
  growth$height,
  k = 15,
  bs = "tp",
  arg = seq(1, 18, l = 80),
  global = TRUE
  # family = gaussian(link = "log") # ensures positivity of velocity
) |>
  tf_derive()
```

#### Data representation matters

The raw finite-difference velocity estimates are jagged (30 irregularly spaced points from the original measurement grid). SRVF registration on such noisy inputs tends to produce wiggly warps that chase noise rather than genuine phase variation. Converting to a smooth representation first gives SRVF cleaner input and much better results.

```{r growth-raw-vs-smooth, eval = have_srvf, include = have_srvf, class.source = "fold-show"}
# SRVF on raw (noisy) velocity:
warp_raw <- tf_register(growth$raw_vel, method = "srvf")
reg_raw <- tf_unwarp(growth$raw_vel, warp_raw)

# SRVF on smooth velocity:
warp_smooth <- tf_register(growth$smooth_vel, method = "srvf")
reg_smooth <- tf_unwarp(growth$smooth_vel, warp_smooth)
```

```{r growth-raw-vs-smooth-plot, eval = have_srvf, include = have_srvf, fig.width = 8, fig.height = 5.5}
reg_brks <- range(c(tf_evaluations(reg_raw), tf_evaluations(reg_smooth))) |>
  (\(x) seq(x[1], x[2], l = 13))()

layout(t(matrix(1:8, 4, 2)))

plot(growth$raw_vel, main = "Raw Velocity (finite diff.)",
     xlab = "Age [years]", ylab = "cm/year", lwd = 0.8, ylim = c(0, 30))
plot(tf_invert(warp_raw), points = FALSE, main = "SRVF Warps (raw)",
     xlab = "Chronological Age", ylab = "Registered Age", lwd = 0.8)
abline(0, 1, lty = 3)
plot(reg_raw, main = "SRVF Registered (raw)",
     xlab = "Registered Age [years]", ylab = "cm/year", lwd = 0.8, ylim = c(0, 30))
plot(reg_raw, main = "SRVF Registered (raw)",
     type = "lasagna", col = hcl.colors(12, rev = TRUE), breaks = reg_brks)

plot(growth$smooth_vel, main = "Smooth Velocity (tfb deriv.)",
     xlab = "Age [years]", ylab = "cm/year", lwd = 0.8, ylim = c(0, 30))
plot(tf_invert(warp_smooth), points = FALSE, main = "SRVF Warps (smooth)",
     xlab = "Chronological Age", ylab = "Registered Age", lwd = 0.8)
abline(0, 1, lty = 3)
plot(reg_smooth, main = "SRVF Registered (smooth)",
     xlab = "Registered Age [years]", ylab = "cm/year", lwd = 0.8, ylim = c(0, 30))
plot(reg_smooth, main = "SRVF Registered (smooth)",
     type = "lasagna", col = hcl.colors(12, rev = TRUE), breaks = reg_brks)
```

The top row shows extreme, biologically implausible warps (i.e. mapping chronological ages 5-8 to "registered" ages < 2) with local wiggles that are more likely to reflect noise in the raw velocity estimates rather than genuine timing differences. The bottom row, using smoother velocity curves, produces simpler and less extreme warps that capture the pubertal timing variation — the main feature of interest.

#### Penalization

Even on noisy inputs, SRVF penalization (`lambda > 0`) can help suppress over-warping. Higher `lambda` pulls warps toward the identity (no warping), trading less strict alignment for smoother and more subtle  warps.

```{r growth-penalization, eval = have_srvf, include = have_srvf, class.source = "fold-show"}
# Penalized SRVF on the raw velocity:
warp_raw_pen <- tf_register(growth$raw_vel, method = "srvf", lambda = 0.1)
reg_raw_pen <- tf_unwarp(growth$raw_vel, warp_raw_pen)
```

```{r growth-penalization-plot, eval = have_srvf, include = have_srvf, fig.width = 8, fig.height = 3}
layout(t(matrix(1:4, 4, 1)))

plot(growth$raw_vel, main = "Raw Velocity",
     xlab = "Age [years]", ylab = "cm/year", lwd = 0.8)
plot(tf_invert(warp_raw_pen), points = FALSE,
     main = expression("Penalized Warps (raw, " * lambda * " = 0.1)"),
     xlab = "Chronological Age", ylab = "Registered Age", lwd = 0.8)
abline(0, 1, lty = 3)
plot(reg_raw_pen,
     main = expression("Penalized Registered (raw, " * lambda * " = 0.1)"),
     xlab = "Registered Age [years]", ylab = "cm/year", lwd = 0.8)
plot(reg_raw_pen, main = "Penalized Registered (raw)",
     type = "lasagna", col = hcl.colors(12, rev = TRUE), breaks = reg_brks)
```

Penalization smooths out the worst warp artifacts, and the result here looks more reasonable than the result for the (oversmoothed) `tfb`-inputs above. 
That said, in most cases the best strategy is likely to be to (carefully!) pre-smooth, and then to use penalization only if the warps still look too flexible.

#### Landmark registration

Landmark registration avoids continuous optimization entirely — it constructs piecewise linear warps from a (user-specified) correspondence of curve features. For growth velocity, natural landmarks are the *end of rapid infant growth* (velocity drops a lot below initial velocity), *pre-pubertal trough* (minimum velocity before the puberty growth spurt), the *pubertal peak* (maximum velocity during puberty). We identify these by searching for the minimum in age 5--(peak-1) and the maximum in age 8--17 on each smooth velocity curve.

```{r growth-landmarks, class.source = "fold-show"}
# End of rapid infant growth: 
#    less than 2/3 of max early childhood growth (1-3) velocity before age 5
growth_slows <- growth$smooth_vel |> tf_zoom(begin = 1, end = 5) |> 
  tf_where(value < 0.66 * max(value[1:10]), return = "first") |> unlist()

# Pubertal peak: maximum velocity in age 8--17
growth_peaks <- growth$smooth_vel |> tf_zoom(begin = 8, end = 17) |> 
  tf_where(value == max(value)) |> unlist()

# Pre-pubertal trough: minimum velocity between age 5 and 1 year before the peak
growth_troughs <- growth$smooth_vel |> tf_zoom(begin = 5, end = growth_peaks - 1) |> 
  tf_where(value == min(value)) |> unlist() 

# Build landmark matrix and register:
growth_lm <- cbind(trough = growth_slows, trough = growth_troughs, peak = growth_peaks)
warp_lm <- tf_register(growth$smooth_vel, method = "landmark", landmarks = growth_lm)
reg_lm <- tf_unwarp(growth$smooth_vel, warp_lm)
```

```{r growth-landmarks-plot, fig.width = 8, fig.height = 3}
layout(t(1:4))
plot(growth$smooth_vel, main = "Smooth Velocity",
     xlab = "Age [years]", ylab = "cm/year", lwd = 0.8)
plot(tf_invert(warp_lm), points = FALSE, main = "Landmark Inverse Warps",
     xlab = "Chronological Age", ylab = "Registered Age", lwd = 0.8)
abline(0, 1, lty = 3)
plot(reg_lm, main = "Landmark Registered",
     xlab = "Registered Age [years]", ylab = "cm/year", lwd = 0.8)
plot(reg_lm, main = "Landmark Registered",
     type = "lasagna", col = hcl.colors(12, rev = TRUE), breaks = reg_brks)
```

The piecewise linear warps align the pubertal peaks well, with the pubertal trough and infant growth landmarks anchoring the earlier phases. Compared to SRVF, the warps are simpler and fully interpretable — each segment directly corresponds to a physiological period (early childhood, late childhood, pubertal acceleration, post-peak deceleration). The trade-off is that landmark registration cannot adapt between landmarks (alignment between successive landmarks depends entirely on those two anchor points) and that landmark definition is often somewhat arbitrary (c.f. definition of "end of rapid infant growth" above...).
