---
title: "`tidyfun`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = TRUE, error = TRUE)
options(digits = 3)
library(devtools)
library(refund)
library(FDboost)
library(tidyverse)
load_all("~/Work/fda/tidyfun")
dti <- with(refund::DTI, data.frame(id = ID, sex = sex))
dti$cca <- feval(DTI$cca, seq(0,1, l = 93), signif = 3)
dti$rcst <- feval(DTI$rcst, seq(0,1, l = 55), signif = 3)
set.seed(1221)
```

# `tidyfun`

The goal of tidyfun is to provide a `tidyverse`-compliant, accessible and well-documented way to handle functional data during data wrangling and exploratory analysis.

```{r, data}
head(dti)
dim(dti)
nrow(dti %>%  filter(rcst[, .7] > .8))

plot(dti$cca, points = FALSE)
lines(mean(dti$cca), col = "red")
lines(mean(dti$cca) + sd(dti$cca), col = "blue", lty = 2)
lines(mean(dti$cca) - sd(dti$cca), col = "blue", lty = 2)

plot(dti$rcst, type = "lasagna")
funplot(dti$rcst, type = "lasagna")
funplot(dti$rcst) + theme_minimal()

# to come:
# dti %>% group_by(sex) %>% mutate(mean_cca = mean(cca), sd_cca = sd(cca))
```


# `fvector`-class

`tidyfun` defines a new type of vector data type for functional data: `fvector`

* basically a `list` of numeric vectors
* with additional attributes that help define *function-like* behavior:
    * mostly how to **evaluate** the functional data on new argument values
    * their domain
    * the resolution of the argument values

## Operations on `fvector`-objects

### Subset & subassign:
```{r}
test <- dti$cca[1:5]
test[1:2] <- test[2:1]
test
```

### Evaluation

```{r}
test[1:2]
test[1:2, seq(0,1, l = 11)]
test[1:2, seq(0,1, l = 11), matrix = FALSE]
test[1:2, seq(0,1, l = 11), interpolate = FALSE]
```

### Compare, compute & summarize:

```{r}
test[1] + test[1] == 2 * test[1]
log(exp(test[2])) == test[2]
test - 1 != test 
mean(test)
sd(test)
median(test) 
# to come: MBD-based median and quantiles
```

## Convert & construct

- convert to & from `list`, `matrix`, 
`data.frame` with columns for `"id"`, `"argvals"`, `"data"`

```{r}
str(m_test <- as.matrix(test))
str(df_test <- as.data.frame(test))
feval(m_test)
feval(df_test)
```

- convert to & from raw functional data to basis representation (more later):

```{r}
fbase(test)
# as.matrix(fbase(test)) 
# as.data.frame(fbase(test))
fbase(m_test)
fbase(df_test)
feval(fbase(test))
```

### Visualize (already shown)

both for `base` and `ggplot2`, either spaghettis (line plots) or lasagna (heat map)

### Derivatives & Integrals: 

*not yet implemented, working on it...*


## `fvector` subclass: `feval` 

Functional data sometimes needs to be kept in raw format: `feval` objects

 - keep a list of `evaluations` and corresponding `argvals`
 - have an associated `evaluator` function that defines how to inter-/extrapolate
 - (internal) subclasses: regular and irregular `feval`

```{r}
x <- seq(-pi, pi, l = 21)
y1 <- sin(2 * x); y2 <- sin(3 * x)
f <- feval(argvals = x, rbind(y1, y2))
f
str(f)

argvals(f)
domain(f)
evaluations(f)
evaluator(f)
f_spline <- f; evaluator(f_spline) <- approx_spline

plot(f, points = FALSE); lines(f_spline, col = "red")
```

##### Irregular:

```{r}
dti$rcst
str(dti$rcst[1:5])
```

## `fvector` subclass: `fbase`

Functional data in basis representation: `fbase` objects

 - keep a list of `coefficients` and a corresponding `basis_matrix` of basis function evaluations
 - have an associated `basis` function that defines how to compute basis for new `argvals`
 - (internal) flavors: `mgcv`'s spline bases and `fpc`-bases (*TODO: wavelet bases*)
 
#### spline-based `fbase`

- accept all arguments of `mgcv`'s `s()`-syntax (univariate only, currently....)
- either do a penalized fit to given data with obs.-specific smoothing or unpenalized (*TODO: global smoothing*)

```{r}
test_b <-fbase(test) 
str(test_b)
```

```{r}
fbase(test, bs = "tp", k = 55)
fbase(test, penalized = FALSE, k = 60)
plot(test, pch = ".", alpha = 1)
lines(fbase(test), col = 2, lty = 3)
lines(fbase(test, penalized = FALSE, k = 60), col = "cyan", lty = 3)
```

#### fpc-based `fbase`

- uses either simple unregularized SVD of the data matrix 
(i.e., not available/sensible for irregular ATM) or `fpca.sc`
- corresponding FPC basis (and mean function) are simply saved as an `feval`-object

```{r}
fpcbase(test, smooth = FALSE, pve = .999)
(test_fpc <- fpcbase(test, pve = .95))
str(test_fpc)

plot(test, pch = ".")
lines(test_fpc, col = 2, lty = 3)
lines(fpcbase(test, smooth = FALSE, pve = .999), col = "cyan", lty = 3)
```

## TODOs: 

  - derivatives: might be fairly easy for `fbase` since `mgcv` offers derivatives of its bases
  - registering/warping should be mostly easy, just overwrite `argvals` (or wrap warping around `evaluator`...?)
  - intensive testing with diverse use-cases
  - extensions for multivariate and image data (will be hard)
  - integration with renovated `refund` for modeling etc.

## ISSUES:

  - lots of `tibblyverse` adjustments still needed (no grouped operations possible ATM, no pretty printing)
  - is `signif_argvals` reasonable?
  - no `S4` means no multiple inheritance for orthogonal implementation of aspects "representation" and "function properties" like monotonous or strictly positive functions in basis or raw data representation.
  - more issues: [https://github.com/fabian-s/tidyfun/issues]
  
